---
title: "Data handling"
output: 
  rmarkdown::html_vignette:
    toc: true
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Data handling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = "../tests/data/TestProject/Code/"
)
```

```{r, echo = FALSE, results = "hide", message = FALSE}
library(esqlabsR)
```

The workflow of modeling and simulation involves running simulations and comparing simulated data against observed data. Functionalities of `esqlabsR` require observed data to be present as [`DataSet`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataSet.html) objects. Please refer to the article [Observed data](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/observed-data.html) for information on how to load data from Excel or *.pkml files.

The function `loadObservedData()` facilitates loading data in standard esqlabs projects. Assuming the standard project folder structure is followed, and Excel files with observed data are present in the `Data` folder, the following chunk of code loads the data:

```{r loadObservedData}
# Create a ProjectConfiguration based on the file located in `tests/data/TestProject/Code`
projectConfiguration <- createDefaultProjectConfiguration()
dataSheets <- "Laskin 1982.Group A"
observedData <- loadObservedData(projectConfiguration = projectConfiguration, sheets = dataSheets)

print(names(observedData))
```
The `loadObservedData()` function loads the data from a path stored in the `projectConfiguration$dataFile`. This function returns a list of `DataSet` class instances. The resulting object will be used later to [run simulations](example-workflow.html), [plot results](example-visualization.html) and [run a sensitivity analysis](example-sensitivity.html).

More detailed information on function signatures can be found in:

* `esqlabsR` documentation on:
    * [createDefaultProjectConfiguration()](https://esqlabs.github.io/esqlabsR/reference/createDefaultProjectConfiguration.html)
    * [ProjectConfiguration class](https://esqlabs.github.io/esqlabsR/reference/ProjectConfiguration.html)
    * [loadObservedData()](https://esqlabs.github.io/esqlabsR/reference/loadObservedData.html)
* `ospsuite` documentation on:
    * [DataImporterConfiguration class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataImporterConfiguration.html)
    * [createImporterConfigurationForFile()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/createImporterConfigurationForFile.html)
    * [loadDataImporterConfiguration()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataImporterConfiguration.html)
    * [DataSet class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataSet.html)
    * [dataSetToDataFrame()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/dataSetToDataFrame.html)
    * [loadDataSetsFromExcel()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataSetsFromExcel.html)
    * [loadDataSetFromPKML()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataSetFromPKML.html)
    * [saveDataSetToPKML()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/saveDataSetToPKML.html)
