---
title: "Data handling"
output: 
  rmarkdown::html_vignette:
    toc: true
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Data handling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE, results = "hide", message = FALSE}
library(esqlabsR)
```

This package utilites the functionalities for reading data from excel files 
implemented in the `ospsuite` package. To read data from excel defined in a
project configuration, use function `loadObservedData()`. The function returns 
a list of `DataSet` objects.

For further information, see `ospsuite` documentation on:

- [DataImporterConfiguration](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataImporterConfiguration.html)
- [createImporterConfigurationForFile()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/createImporterConfigurationForFile.html)
- [loadDataImporterConfiguration()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataImporterConfiguration.html)
- [DataSet](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataSet.html)
- [dataSetToDataFrame](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/dataSetToDataFrame.html)
- [loadDataSetsFromExcel()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataSetsFromExcel.html)
- [loadDataSetFromPKML()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataSetFromPKML.html)
- [saveDataSetToPKML()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/saveDataSetToPKML.html)
