---
title: "Design scenarios"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Design scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

A **Scenario** is the unit of work in `esqlabsR`: one runnable simulation, described declaratively. It names the model to run, the subject to run it on, the dosing, the parameters to apply, the outputs to record, and the time grid. You author scenarios (and the definitions they reference) into a Project, then hand the Project to `runScenarios()` to execute them. This article shows how to build scenarios and their supporting definitions from R, and how to edit, rename, duplicate, and remove them.

This article assumes you know what a Project is and how to load one; see `vignette("projects")` if not.

## Working against a writable copy

Every authoring function in this article mutates a Project, and scenario edits are **write-through**: the moment you add or change a scenario, its definition file is written to disk under `definitions/scenarios/`. So you never want to author against the package's own bundled example. Instead, scaffold a fresh, writable copy of the worked example into a temporary directory and load that:

```{r}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))
project
```

In your own work you would point `loadProject()` at your real project instead; everything below operates on this `project` object. Each definition kind is reachable as a named list on the project, keyed by id, so you can inspect what is defined at any time, for example `project$scenarios`:

```{r}
project$scenarios
```

## Building the supporting definitions

A scenario can only reference definitions that already exist, so before you create the scenario you typically define the pieces it will use. The authoring API is uniform: every function takes the project first, then the definition's `id`, then its fields.

Because an id can become a filename in the `definitions/` tree, ids are **canonicalized** the moment they enter the project: lowercased and made filename-safe. When canonicalization changes the id you typed, the authoring function warns you and stores the canonical form, as the `addOutputPath()` call below shows. See `vignette("projects")` for the full treatment.

### Output paths

An Output Path maps a short id to a full model path in OSPS notation. Recording results by id keeps scenario definitions readable and lets several scenarios share the same output:

```{r, warning = TRUE}
addOutputPath(
  project,
  id = "Aciclovir/Liver:Conc",
  path = "Organism|Liver|Intracellular|Aciclovir|Concentration in container"
)
project$outputPaths[["aciclovir_liver_conc"]]
```

The id you typed was stored as `aciclovir_liver_conc`. Where the id is already canonical no warning appears:

```{r}
addOutputPath(
  project,
  id = "aciclovir_kidney",
  path = "Organism|Kidney|Intracellular|Aciclovir|Concentration in container"
)
```

### Parameter sets

Model parameters are organized into named **Parameter Sets**. A set is a group of parameter entries that you apply together. You create an empty set with `addParameterSet()` and fill it with `addParameterEntry()`, one entry at a time. Each entry is a `(containerPath, parameterName, value, units)` quadruple:

```{r}
addParameterSet(project, "ckd")
addParameterEntry(
  project,
  "ckd",
  containerPath = "Organism|Kidney",
  parameterName = "GFR",
  value = 30,
  units = "ml/min"
)
project$parameterSets[["ckd"]]
```

Any definition (scenario, individual, or application) references parameter sets from this one collection by id, so you can compose a shared `global` set with disease-specific sets like `ckd` per scenario. `addParameterEntry()` is last-write-wins on a repeated `(containerPath, parameterName)` pair, and creates the set on demand if it does not yet exist.

### Individuals

An **Individual** carries the biometrics of a single simulated subject plus any individual-specific parameter sets. `species` and `gender` are required; the rest are optional named fields:

```{r}
addIndividual(
  project,
  "adult_female",
  species = "Human",
  gender = "FEMALE",
  weight = 60,
  height = 165,
  age = 35,
  parameterSets = "adult_male_default"
)
project$individuals[["adult_female"]]
```

The `parameterSets` argument is a character vector of ids referencing the project's parameter sets; it is validated immediately, so a typo in a set id is caught at add time rather than at run time.

The authoring functions are **vectorized over the id**: pass a vector of ids to add several definitions in one call (one write-through). The id sets the count, a scalar field is recycled to all of them, a length-matched vector is aligned by position, and a field that is itself a vector for one definition (such as `parameterSets`) is applied whole to each. The call is all-or-nothing: if any one definition fails validation, nothing is written.

```{r}
addIndividual(
  project,
  c("teen_male", "teen_female"),
  species = "Human",
  gender = c("MALE", "FEMALE"),
  weight = c(50, 48),
  height = 160,
  age = 14,
  parameterSets = "adult_male_default"
)
names(project$individuals)
```

### Populations

A **Population** describes a virtual population to sample: a species, a number of individuals, and optional demographic ranges (age, weight, height, BMI) and their units. A scenario that references a population becomes a population simulation:

```{r}
addPopulation(
  project,
  "pediatric",
  species = "Human",
  numberOfIndividuals = 25,
  ageMin = 2,
  ageMax = 12,
  proportionOfFemales = 50
)
project$populations[["pediatric"]]
```

### Applications

An **Application** describes how the substance is administered, stored as a reference to the parameter set that holds its dose, start time, and other parameters:

```{r}
addApplication(
  project,
  "aciclovir_iv_500mg",
  parameterSets = "aciclovir_iv_250mg_default"
)
project$applications[["aciclovir_iv_500mg"]]
```

## Creating a scenario

With the supporting definitions in place, `addScenario()` assembles them into a runnable scenario. The id comes first, right after the project, and `modelFile` is the only other required argument; everything else is an optional reference or setting:

```{r}
addScenario(
  project,
  "aciclovir_iv_ckd",
  modelFile = "Aciclovir.pkml",
  individual = "adult_male",
  application = "aciclovir_iv_250mg",
  parameterSets = c("global", "aciclovir", "ckd"),
  outputPaths = c("aciclovir_pvb", "aciclovir_kidney"),
  simulationTime = "0, 24, 60",
  simulationTimeUnit = "h"
)
project$scenarios[["aciclovir_iv_ckd"]]
```

A scenario ties together several pieces, all referenced by id and all optional except the model file:

- **`modelFile`** (required): the `.pkml` file holding the model structure, named relative to the project's model folder.
- **`individual`**: a named Individual applied before simulating. Omit it to simulate the subject as exported from PK-Sim or MoBi.
- **`population`**: a Population to sample. Supplying one makes the scenario a population simulation.
- **`application`**: an Application describing how the substance is administered.
- **`parameterSets`**: Parameter Sets layered onto the model, in the order you list them (later sets override earlier ones).
- **`outputPaths`**: the Output Paths to record.
- **`simulationTime`** / **`simulationTimeUnit`**: the simulated time grid and its unit.
- **`steadyState`**: whether to pre-run the model to a steady state first.

Two behaviors matter here:

- Every reference is checked as the scenario is added. If `individual`, `application`, or any entry in `parameterSets` does not resolve to something defined in the project, `addScenario()` aborts and the scenario is not created. This is why you build the supporting definitions first.
- `parameterSets` is applied in order: `global` first, then `aciclovir`, then `ckd`, with later sets overriding earlier ones where they touch the same parameter.

The `simulationTime` argument is a string of one or more intervals. A single interval is `"start, end, resolution"`, where resolution is the number of points per unit of `simulationTimeUnit`. Several intervals are separated by `;`, for example `"0, 20, 60; 20, 504, 1"` to simulate the first 20 hours densely and a longer tail coarsely.

To make it a population scenario, pass a `population` instead of (or alongside) an `individual`; the scenario's type switches to a population simulation automatically. For a steady-state run, set `steadyState = TRUE` and give `steadyStateTime` and `steadyStateTimeUnit`. As with every authoring call, the scenario is written through to `definitions/scenarios/aciclovir_iv_ckd.json` the moment it is added.

## Editing a scenario in place

You will rarely get a scenario right on the first try. Rather than removing and re-adding it, edit it in place with `setScenario()`. It mirrors `addScenario()` exactly, but with two important behaviors:

- **Only the fields you pass change.** Every argument you omit keeps its current value (a partial update).
- **Passing `NULL` clears an optional field**, whereas omitting the argument leaves it untouched. (The required `modelFile` cannot be cleared.)

Here we widen the time grid and swap the dose, leaving everything else as it was:

```{r}
setScenario(
  project,
  "aciclovir_iv_ckd",
  application = "aciclovir_iv_500mg",
  simulationTime = "0, 48, 60"
)
project$scenarios[["aciclovir_iv_ckd"]]
```

And here we detach the individual entirely by passing `NULL`, so the scenario simulates the model's own default individual:

```{r}
setScenario(project, "aciclovir_iv_ckd", individual = NULL)
project$scenarios[["aciclovir_iv_ckd"]]$individualId
```

The same partial-update, `NULL`-clears pattern applies to the supporting definitions. `setIndividual()` and `setPopulation()` change named fields on an existing individual or population, and `setOutputPath()` changes the literal path bound to an output-path id (keeping the id, so every scenario that records it keeps working):

```{r}
setIndividual(project, "adult_female", weight = 62, age = 36)
project$individuals[["adult_female"]]

setOutputPath(
  project,
  "aciclovir_kidney",
  path = "Organism|Kidney|Urine|Aciclovir|Concentration"
)
project$outputPaths[["aciclovir_kidney"]]
```

As with `setScenario()`, a reference argument is validated before anything is written, so an invalid edit touches neither memory nor disk. To take a parameter entry back out use `removeParameterEntry()` (the set is auto-removed if it becomes empty), and `removeParameterSet()` drops a whole set.

## Renaming, duplicating, and removing scenarios

A common authoring move is to take a working scenario as a template for a variant. `duplicateScenario()` makes an independent copy under a new id, leaving the original untouched:

```{r}
duplicateScenario(project, "aciclovir_iv", "aciclovir_iv_highdose")
names(project$scenarios)
```

`renameScenario()` changes a scenario's id, moving its definition file accordingly and keeping the stored name in step with the new key:

```{r}
renameScenario(project, "aciclovir_iv_highdose", "aciclovir_iv_high")
names(project$scenarios)
```

Both take the existing id first and the new id second, and both canonicalize their ids the same way `addScenario()` does. Neither will silently overwrite an existing scenario: a rename or duplicate onto an id already in use aborts.

When a scenario has served its purpose, `removeScenario()` drops it from the project and deletes its definition file:

```{r}
removeScenario(project, "aciclovir_iv_high")
names(project$scenarios)
```

## Seeding scenarios from a model file

When you already have one or more `.pkml` files and want a scenario per file, `createScenariosFromPKML()` saves you from writing each `addScenario()` call by hand. It reads each model file, extracts the output selections and the simulation time grid embedded in it, and adds a scenario to the project for each one:

```{r}
pkml_path <- file.path(project_dir, "Models", "Simulations", "Aciclovir.pkml")
createScenariosFromPKML(
  pkmlFilePaths = pkml_path,
  project = project,
  scenarios = "aciclovir_from_pkml"
)
project$scenarios[["aciclovir_from_pkml"]]
```

For each model, the call:

- Registers the output paths found in the model, reusing an existing id when the literal path is already known and generating a readable id otherwise.
- Makes the new scenario names unique against the scenarios already present.
- Recycles a single per-scenario setting (individual, population, application, parameter sets) to every scenario, or aligns a longer vector by position.

Pass a vector of `pkmlFilePaths` to create several scenarios at once. Pass `scenarios` to name them yourself; otherwise the simulation names stored inside the models are used.

This is a starting point, not the finished article: the scenarios it creates are ordinary scenarios, so you refine them afterward with `setScenario()`, attach an individual or parameter sets, and adjust the time grid, exactly as in the sections above.

## Where to go next

You now have the full authoring loop: build the supporting definitions (output paths, parameter sets, individuals, populations, applications), assemble them into scenarios with `addScenario()`, refine them with `setScenario()` and the other field setters, and reshape the set with `duplicateScenario()`, `renameScenario()`, and `removeScenario()`. Every edit is written through to the `definitions/` tree as you make it.

From here, validate what you have built with `validateProject()` (see `vignette("validate-project")`), then run it: `vignette("run-simulations")` covers passing the project to `runScenarios()` and collecting results, and `vignette("plot-results")` covers turning those results into figures.
