---
title: "Sensitivity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{esqlabsR-sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  fig.showtext = TRUE
)
```

```{r setup}
library(esqlabsR)
```

```{r, echo = FALSE, results = "hide", message = FALSE}
library(esqlabsR)
projectConfiguration <- createProjectConfiguration(exampleProjectConfigurationPath())
dataSheets <- "Laskin 1982.Group A"
observedData <- loadObservedData(projectConfiguration = projectConfiguration, sheets = dataSheets)
# Create `ScenarioConfiguration` objects from excel files
scenarioConfigurations <- readScenarioConfigurationFromExcel(
  scenarioNames = "TestScenario",
  projectConfiguration = projectConfiguration
)

scenarios <- createScenarios(scenarioConfigurations = scenarioConfigurations)

simulatedScenarios <- runScenarios(
  scenarios = scenarios
)
```

### Workflow

Sensitivity analysis quantifies how the pharmacokinetics of the drug
changes with a variation of simulation parameters. This is important to
track if the values of simulation parameters are uncertain. Further
information about sensitivity analysis and the mathematical background
is provided in the OSPS documentation section [Sensitivity
Analysis](https://docs.open-systems-pharmacology.org/shared-tools-and-example-workflows/sensitivity-analysis).

In the aciclovir simulation example, the lipophilicity of aciclovir was
assumed to be -0.097 in log units. In the sensitivity analysis, we want
to quantify how much the pharmacokinetic parameters change if the
lipophilicity of aciclovir is varied by certain factors.

The `sensitivityCalculation()` function in the `{esqlabsR}` package does
that by re-running the simulation with a set of updated parameter
values. By default, the specified parameter will be multiplied by 0.1,
0.2, 0.3, ..., 1, 2, 3, ... and 10, and for each value a simulation will
be run. These factors can be customized by the `variationRange`
argument. The function returns a list with output paths for which the
sensitivity has been calculated, paths of parameters that have been
varied, a `SimulationResults` object, and a data frame with calculated
PK parameters for each of the input parameter values.

```{r analysis}
simulation <- simulatedScenarios$TestScenario$simulation
outputPaths <- enum(list(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
))
parameterPaths <- c(
  "Lipophilicity" = "Aciclovir|Lipophilicity",
  "Dose" = "Applications|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  "GFR fraction" = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Glomerular Filtration-GFR|GFR fraction"
)
analysis <- sensitivityCalculation(simulation, outputPaths, parameterPaths)
head(analysis$pkData)
```

In the aciclovir example case, the default value of lipophilicity is
-0.097 log units, corresponding to the area under the curve (AUC) of
3915.87 µmol×min/L. Increasing the lipophilicity to -0.0097 log units
leads to a decrease of AUC to 3895.43 µmol×min/L (a change of -0.52%),
while decreasing the lipophilicity to -0.97 log units leads to an
increase of AUC to 4015.73 µmol×min/L (a change of 2.55%). The
sensitivity of AUC to 10-fold increase of lipophilicity is calculated as
$$\frac{\Delta AUC}{\Delta lipophilicity} \times \frac{lipophilicity}{AUC}=\frac{4015.73-3915.87}{-0.97--0.097} \times \frac{-0.97}{4015.73}=0.027$$
```{r}
analysis$pkData |>
  dplyr::filter(
    ParameterPath == "Aciclovir|Lipophilicity",
    PKParameter == "AUC_inf",
    ParameterFactor %in% c(0.1, 1, 10)
  ) |>
  dplyr::select(ParameterFactor, PKParameterValue, 
                PKPercentChange, SensitivityPKParameter) |>
  dplyr::mutate(
    PKPercentChange = round(PKPercentChange, 2),
    SensitivityPKParameter = round(SensitivityPKParameter, 4)
  )
```

The results of the sensitivity analysis can be plotted using the following 
functions:

-   the `sensitivitySpiderPlot()` function generate a separate plot for
    each of the PK parameters under investigation. By default, the
    `sensitivityCalculation()` function computes the changes in area
    under the curve (`AUC_inf`), maximum concentration (`C_max`), and
    time when the maximum concentration is reached (`t_max`).

```{r spider-plot, fig.width=7, fig.height=4}
sensitivitySpiderPlot(analysis)
```

-   the `sensitivityTimeProfiles()` function plots the concentration
    profile for each of the input parameter values:

```{r time-profile, fig.width=7, fig.height=4}
sensitivityTimeProfiles(analysis)
```

- the `sensitivityTornadoPlot` function visualizes the impact of increasing or decreasing input parameters on PK parameters at a fixed scaling factor. The scaling factor to plot (default: 0.1) and its reciprocal (e.g., 10) must both be included in the sensitivity analysis result.
While only a single parameter (lipophilicity) is varied in this example, tornado plots are especially useful for side-by-side comparisons when multiple parameters are analyzed.

```{r tornado-plot, fig.width=7, fig.height=3.5}
sensitivityTornadoPlot(analysis)
```


### Sensitivity calculation for user-defined (non-PK parameters) outputs

Though the typical workflows for sensitivity analysis are focused on PK
parameters, it is also possible to calculate the sensitivities for any
numerical model outputs using user-defined functions. As for the PK
parameters, the provided function(s) will be applied to each output
defined in the argument `outputPaths` of the `sensitivityCalculation()`
function.

The custom functions are provided as a named list of functions in the
`customOutputFunctions` argument. Each function must have the arguments
`x` and `y`, through which it accesses the simulated time values (`x`)
or the output values (`y`). The function should return a single numeric
value.

In the following example, we calculate the sensitivity of the average
concentration of aciclovir in the peripheral venous blood to the
lipophilicity of aciclovir.

We first define a function that calculates the mean of a given numerical
vector:

```{r custom-function}
meanFunction <- function(x, y) {
  mean(y)
}
```

Next, we run the sensitivity analysis for the average concentration of
aciclovir in the peripheral venous blood:

To define custom labels for the parameters in the resulting plots, you can pass 
a *named vector* to the `parameterPaths` argument, where the names will be used 
as labels instead of full paths.

```{r custom-analysis}
simulation <- simulatedScenarios$TestScenario$simulation
customOutputPaths <- enum(list(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
))
customParameterPaths <- c("Lipophilicity" = "Aciclovir|Lipophilicity")

customAnalysis <- sensitivityCalculation(
  simulation,
  customOutputPaths,
  pkParameters = list(),
  parameterPaths = customParameterPaths,
  customOutputFunctions = list(AverageConcentration = meanFunction)
)
head(customAnalysis$pkData)
```

[**Note**]{.underline}: If we want to calculate the sensitivity of the
custom function only without the default PK-parameters, we need to set
the `pkParameters` value to an empty list.

Finally, we plot the sensitivity of the average concentration of
aciclovir in the peripheral venous blood to the lipophilicity of
aciclovir. 
```{r custom-plot, fig.width=7, fig.height=4}
sensitivitySpiderPlot(customAnalysis)
```

### Troubleshooting

-   The `SensitivityPKParameter` column in the output of
    `analysis$pkData` will be `NA` in the rows corresponding to the
    initial parameter values (i.e., where the multiplication factor is
    1).

More detailed information on function signatures can be found in:

-   `esqlabsR` documentation on:
    -   [sensitivityCalculation()](https://esqlabs.github.io/esqlabsR/reference/sensitivityCalculation.html)
    -   [sensitivitySpiderPlot()](https://esqlabs.github.io/esqlabsR/reference/sensitivitySpiderPlot.html)
    -   [sensitivityTimeProfiles()](https://esqlabs.github.io/esqlabsR/reference/sensitivityTimeProfiles.html)
