---
title: "Get Started"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to esqlabsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

## Introduction

`{esqlabsR}` is built around a single configuration file — `Project.json` — that
defines everything your project needs: where files are stored, which scenarios to
run, what observed data to load, and how to generate plots. Loading it with
`loadProject()` gives you a `Project` object that all other functions operate on.

To follow along with this tutorial, load the example project included in the
package:

```{r}
projectConfigPath <- system.file(
  "extdata", "projects", "Example", "Project.json",
  package = "esqlabsR"
)
myProject <- loadProject(projectConfigPath)
```

## Project

`Project.json` is the single source of truth for your project. It opens with a
`filePaths` block that tells `{esqlabsR}` where to find every supporting file:

```json
"filePaths": {
  "modelFolder": "Models/Simulations/",
  "configurationsFolder": "Configurations/",
  "modelParamsFile": "ModelParameters.xlsx",
  "individualsFile": "Individuals.xlsx",
  "populationsFile": "Populations.xlsx",
  "applicationsFile": "Applications.xlsx",
  "plotsFile": "Plots.xlsx",
  "dataFolder": "Data/",
  "outputFolder": "Results/"
  // ... (additional keys omitted)
}
```

`loadProject()` reads these paths and resolves them relative to the JSON file's
location. Printing the resulting `Project` object shows all resolved paths.

For full details on the folder layout, validation, and version control workflows,
see `vignette("project-structure")`.

## Scenarios

A scenario is the core unit of simulation. Each scenario entry in the `scenarios`
array links together a model file, parameter groups, individual or population
settings, application protocol, and simulation time window:

```json
{
  "name": "Aciclovir_iv",
  "modelFile": "Aciclovir.pkml",
  "individualId": "Adult_male",
  "modelParameterGroups": ["Global", "Aciclovir"],
  "applicationProtocol": "Aciclovir_iv_250mg",
  "simulationTime": "0, 24, 60",
  "simulationTimeUnit": "h"
}
```

To run scenarios, pass the `Project` object to `runScenarios()`:

```{r, eval=FALSE}
myScenarioResults <- runScenarios(myProject, scenarioNames = "Aciclovir_iv")
```

`runScenarios()` returns a named list — results for each scenario are accessible
by name, e.g. `myScenarioResults$Aciclovir_iv`.

For a full description of all scenario parameters (steady-state, population
simulations, output paths, etc.), see `vignette("design-scenarios")` and
`vignette("run-simulations")`.

## Data

Observed data is declared in the `observedData` section of `Project.json` and
loaded automatically when you call `loadProject()` — no separate import step
needed:

```json
"observedData": [
  {
    "type": "excel",
    "sheets": ["Laskin 1982.Group A"]
  }
]
```

The Excel file that is read is the one declared under `dataFile` in the `filePaths` section of `Project.json`.
The loaded datasets are available via `myProject$observedData`, a named list of
`DataSet` objects. When plotting, observed data is combined with simulation
results using a `DataCombined` object — the `plots` section of `Project.json`
handles this automatically. See `vignette("plot-results")` for details on
`DataCombined` and data transformations.

## Plots

Plots are fully driven by the `plots` section of `Project.json`. It contains
three sub-sections:

- **`dataCombined`** — which simulation outputs and observed datasets to combine,
- **`plotConfiguration`** — what type of plot to generate for each data combined,
- **`plotGrids`** — how to arrange plots into figures and export them.

A `plotConfiguration` entry looks like this:

```json
{
  "plotID": "P1",
  "DataCombinedName": "Aciclovir_individual",
  "plotType": "individual",
  "title": "Aciclovir IV 250 mg"
}
```

To generate all plots defined in the JSON, call `createPlots()`:

```{r, echo=FALSE}
myScenarioResults <- runScenarios(myProject, scenarioNames = c("Aciclovir_iv", "Aciclovir_iv_female"))
```

```{r, out.width="100%", fig.align="center"}
plots <- createPlots(myProject, simulatedScenarios = myScenarioResults)
plots$Individual_diagnostics
```

`createPlots()` returns a named list of plot grid objects — one per `plotGrids`
entry. If `exportConfiguration` is defined in the JSON, figures are also written
to disk in `Results/Figures/`.

For the full schema of all plot types, grid layout options, and export
configuration, see `vignette("plot-results")`.

## Full Workflow

Here is the complete end-to-end workflow:

```{r, eval=FALSE}
library(esqlabsR)

# Load the project
projectConfigPath <- system.file(
  "extdata", "projects", "Example", "Project.json",
  package = "esqlabsR"
)
myProject <- loadProject(projectConfigPath)

# Run scenarios (omit scenarioNames to run all scenarios defined in Project.json)
myScenarioResults <- runScenarios(myProject, scenarioNames = "Aciclovir_iv")

# Generate and export all plots
createPlots(myProject, simulatedScenarios = myScenarioResults)
```
