---
title: "Introduction to ESQlabsR"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to ESQlabsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
```

`esqlabsR` builds modeling and simulation workflows on top of the Open Systems Pharmacology Suite (OSPS). It organizes everything a PBPK or QSP analysis needs (the model, the subjects you simulate, the dosing, the parameters you apply, the outputs you record, the observed data you compare against, and the figures you produce) into a single **Project**. You load a Project once, then run scenarios and plot results from it.

This tutorial walks the whole workflow once, against the example project the package ships. You will load a project, look inside it, add an individual and a scenario, describe a figure, then run the scenario and draw the figure. It takes a few minutes, every step runs as written, and nothing you do here touches your own files.

You do not need to understand each piece as you go. The point is to see the shape of the loop; the how-to guides fill in the detail afterwards.

## Install and load

`esqlabsR` depends on the OSPS R packages. Install those via the Open Systems Pharmacology setup instructions, then install and load `esqlabsR`:

```{r}
library(esqlabsR)
```

## Load a project

The package bundles a complete example Project. Because you will save changes to it later, take your own copy first: `initProject()` scaffolds one into a temporary directory, and `loadProject()` reads it into memory. (`initProject()` takes the directory; `loadProject()` takes the `Project.json` file inside it.)

```{r}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))
```

Printing the Project gives you a quick map of its name, location, and definition counts.

```{r}
print(project)
```

## Look inside the project

A project is built out of **definitions**: named pieces such as a scenario, an individual, or a plot. Each kind is a named list on the project, keyed by id. Printing one shows how many there are and what they are called:

```{r}
project$definitions$individuals
```

Printing a single definition shows its configured fields, so you can inspect one without opening a file:

```{r}
project$definitions$individuals$adult_male
```

That is enough to work with for now. (`vignette("reference-project-file")` describes how the pieces are stored on disk, if you want it later.)

## Author a scenario and a figure

The authoring API is uniform: every function takes the `project` first, then the definition's `id`, then its fields. The example only defines an adult male, so we will add an adult female, run aciclovir in her, and plot the result.

Start with a new **Individual**, the subject a scenario simulates.

```{r}
addIndividual(
  project,
  id = "adult_female",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "FEMALE",
  weight = 60,
  height = 165,
  age = 30
)
```

Nothing has been written to disk yet: the new individual is held in memory until you save, which we do in a moment.

With the individual in place, add a **Scenario** that simulates aciclovir in her. A Scenario pins a model to a subject, dosing, parameters, a time grid, and the outputs to record. Only `modelFile` is required; everything else is an optional reference to another definition in the project. References are checked as the scenario is added, so a typo in an id stops you here rather than at run time.

```{r}
addScenario(
  project,
  id = "aciclovir_iv_female",
  modelFile = "Aciclovir.pkml",
  individual = "adult_female",
  application = "aciclovir_iv_250mg",
  parameterSets = c("global", "aciclovir"),
  outputPaths = "aciclovir_pvb",
  simulationTime = "0, 24, 60",
  simulationTimeUnit = "h"
)
```

Finally, describe a figure. A **DataCombined** says which curves the figure draws, here just one: the plasma concentration from the new scenario. A **Plot** then draws a chart from it.

```{r}
addDataCombined(
  project,
  id = "aciclovir_female_combined",
  simulated = list(
    list(
      label = "Aciclovir (female)",
      scenario = "aciclovir_iv_female",
      path = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
    )
  )
)

addPlot(
  project,
  id = "aciclovir_female_profile",
  dataCombined = "aciclovir_female_combined",
  plotType = "individual",
  title = "Aciclovir plasma profile (adult female)"
)
```

Everything so far has changed the project in memory only. `saveProject()` writes the new individual, scenario, DataCombined, and plot to disk:

```{r}
saveProject(project)
```

## Run the scenario

With the scenario defined, run it. `runScenarios()` runs one or more scenarios from the Project and returns a named list keyed by scenario name. We pass the single scenario we just built; the first run of a session initializes PK-Sim and takes a few seconds, so a one-scenario example keeps things quick.

```{r}
res <- runScenarios(project, scenarios = "aciclovir_iv_female")
```

Each entry of the returned list is a **Scenario Result**, bundling the simulation and its outputs. This is what `createPlots()` reads from next.

## Plot the result

`createPlots()` reads the plot definitions stored in the Project and draws them against the scenario results you pass in. Ask for the plot you just authored, by its id:

```{r}
plots <- createPlots(
  project,
  plots = "aciclovir_female_profile",
  scenarioResults = res
)
```

Each entry is an ordinary `ggplot` object, so printing it draws it:

```{r, out.width = "80%", fig.align = "center", fig.width = 6, fig.asp = 0.7, dpi = 150}
plots$aciclovir_female_profile
```

That is the whole loop. You added a subject and a scenario to a project, described a figure, ran the simulation, and drew the figure, all from one object.

## Where to go next

You have seen the shape of the workflow. From here, pick the guide for whatever you need to do next:

- `vignette("how-to-set-up-a-project")` to create your own project, save your work, and share it as a single file.
- `vignette("how-to-design-scenarios")` to author the definitions a scenario is built from, edit them in place, and add several at once.
- `vignette("how-to-validate-a-project")` to check a project hangs together before running it.
- `vignette("how-to-run-simulations")` to run many scenarios, populations, and steady-state runs, and to save and reload results.
- `vignette("how-to-plot-results")` for richer figures, observed data overlays, population percentile bands, and the esqLABS house style.
- `vignette("how-to-use-observed-data")` for the four ways measured data enters a project.

Two other kinds of article are worth knowing about. `vignette("about-the-project-model")` and `vignette("about-the-plotting-model")` explain why the package works the way it does, which is the fastest way to stop being surprised by it. `vignette("reference-definition-fields")` and `vignette("reference-project-file")` are for looking things up: every field of every definition, whether it is required, and what happens when you leave it out.
