---
title: "4. Plot Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{esqlabsR-plot-results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Plotting the simulation results is an important part of model diagnostics and 
quality control. Simulated modeling scenarios can be passed to plotting functions 
from the `{ospsuite}` package to create uniformly-looking plots. To get familiar 
with the `DataCombined` class used to store matching observed and simulated data, 
read the [Working with `DataCombined` class](https://www.open-systems-pharmacology.org/OSPSuite-R/dev/articles/data-combined.html) 
article. The article [Visualizations with `DataCombined`](https://www.open-systems-pharmacology.org/OSPSuite-R/dev/articles/data-combined-plotting.html) 
covers the basics on how to create supported plot types and how to customize them. 

For the following examples, we will simulate an example scenario as described in 
[Standard workflow](standard-workflow.html) and load the corresponding observed 
data as described in [Data handling](data-handling.html).

```{r, echo = TRUE, message = FALSE}
library(esqlabsR)
# Create a project configuration
projectConfiguration <- createDefaultProjectConfiguration(example_ProjectConfiguration())
# Create `ScenarioConfiguration` objects from excel files
scenarioConfigurations <- readScenarioConfigurationFromExcel(
  scenarioNames = "TestScenario",
  projectConfiguration = projectConfiguration
)
# Run scenario configuration
scenarios <- createScenarios(scenarioConfigurations = scenarioConfigurations)
simulatedScenarios <- runScenarios(
  scenarios = scenarios
)

# Load observed data
dataSheets <- "Laskin 1982.Group A"
observedData <- loadObservedData(projectConfiguration = projectConfiguration, sheets = dataSheets)
```

## Create a `DataCombined` object

The simulation results are stored in a list returned by the `runScenarios()` 
function. Plotting and visualization is performed by storing these results 
along with matching observed data in a `DataCombined` object and passing it to 
plotting functions. Observed data in the form of `DataSet` objects are added to a 
`DataCombined` object via the `addDataSets()` function, simulated results can be 
added by using the `addSimulationResults()` function. Observed and simulated 
data can be linked by setting the `groups` argument in both methods. Data sets of the 
same group will then be plotted together when calling plotting functions on the `DataCombined` object.

Let's create a `DataCombined` object and populate it with data with the following code:

```{r datacombined}
dataCombined <- DataCombined$new()
dataCombined$addDataSets(observedData, names = "Observed", groups = "Aciclovir")
dataCombined$addSimulationResults(simulatedScenarios$TestScenario$results,
  names = "Simulated",
  groups = "Aciclovir"
)
```

## Customize and generate plots

Customization of the generated figures - specifying title, axes ranges, axes units, 
position of the legend, etc., are done through *plot configurations* - objects 
of the class [`DefaultPlotConfiguration`](https://www.open-systems-pharmacology.org/OSPSuite-R/dev/reference/DefaultPlotConfiguration.html). 
To combine multiple plots into a multi-panel figure, create a `PlotGridConfiguration` 
object, add plots to it, and plot with the `plotGrid()` method. Finally, to export 
a plot to a file (e.g., `PNG` or `PDF`), use an `ExportConfiguration` object.

To use configurations with similar look and feel in the different `esqLABS` projects, 
create the configurations using the following functions:

  - [`createEsqlabsPlotConfiguration()`](https://esqlabs.github.io/esqlabsR/reference/createEsqlabsPlotConfiguration.html)
  - [`createEsqlabsPlotGridConfiguration()`](https://esqlabs.github.io/esqlabsR/reference/createEsqlabsPlotGridConfiguration.html)
  - [`createEsqlabsExportConfiguration(projectConfiguration)`](https://esqlabs.github.io/esqlabsR/reference/createEsqlabsExportConfiguration.html)
  
For the list of supported properties of the `PlotGirdConfiguration`, refer to the 
[reference](https://www.open-systems-pharmacology.org/TLF-Library/reference/ExportConfiguration.html) 

The next example shows how to create a multi-panel figure using the default 
configurations.

```{r plot-time-profile, fig.width=7, fig.height=7}
plotConfig <- createEsqlabsPlotConfiguration()
gridConfig <- createEsqlabsPlotGridConfiguration()

plotConfig$title <- "Time profile"
indivPlot <- plotIndividualTimeProfile(dataCombined, defaultPlotConfiguration = plotConfig)

plotConfig$title <- "Observed-vs-simulated"
obsVsSimPlot <- plotObservedVsSimulated(dataCombined, defaultPlotConfiguration = plotConfig)

plotConfig$title <- "Res-vs-time"
resVsTimePlot <- plotResidualsVsTime(dataCombined, defaultPlotConfiguration = plotConfig)

plotConfig$title <- "Res-vs-simulated"
resVsSimPlot <- plotResidualsVsSimulated(dataCombined, defaultPlotConfiguration = plotConfig)

gridConfig$addPlots(list(indivPlot, obsVsSimPlot, resVsTimePlot, resVsSimPlot))
gridConfig$title <- "All aciclovir plots"
gridPlot <- plotGrid(gridConfig)
gridPlot
```

## Save Plots

To save the plot to a `PNG` file, use the [`ExportConfiguration`](https://www.open-systems-pharmacology.org/TLF-Library/reference/ExportConfiguration.html). Make sure that `fileName` argument end with 
`.png`:

```{r saveToPng, eval = FALSE}
exportConfig <- createEsqlabsExportConfiguration(projectConfiguration)
exportConfig$savePlot(gridPlot, fileName = "All plots.png")
```

## Details


### Observed Data


The workflow of modeling and simulation involves running simulations and comparing simulated data against observed data. Functionalities of `esqlabsR` require observed data to be present as [`DataSet`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataSet.html) objects. Please refer to the article [Observed data](https://www.open-systems-pharmacology.org/OSPSuite-R/dev/articles/observed-data.html) for information on how to load data from Excel or `*.pkml` files.

`{esqlabsR}` offers a convenience function `loadObservedData()` that facilitates loading data in standard *esqLABS* projects. Assuming the standard project folder structure is followed and a valid `ProjectConfiguration` (see [Standard workflow](standard-workflow.html)) and Excel files with observed data are present in the `projectConfiguration$dataFolder` folder, the following code loads the data:

```{r loadObservedData}
projectConfiguration <- createDefaultProjectConfiguration(example_ProjectConfiguration())
dataSheets <- "Laskin 1982.Group A"
observedData <- loadObservedData(
  projectConfiguration = projectConfiguration,
  sheets = dataSheets
)

print(names(observedData))
```

The function loads the data from the file `projectConfiguration$dataFile` from the folder `projectConfiguration$dataFolder` and returns a list of `DataSet` objects. The resulting object will be used later to [plot results](example-visualization.html) to compare simulated and observed data.


- The `sheets` argument of the `loadObservedData()` function should be a string or a list of strings. If a specified sheet is not found in the file, it will be omitted with a warning; the `observedData` variable may then be an empty named list.

- If the data file specified in `projectConfiguration$dataFile` is missing in the filesystem, the `loadObservedData()` function will fail with an `Invalid File` message.

### Troubleshooting

* At any time, you can check the groups assigned to the datasets in the `DataCombined` 
object by calling the `DataCombined$groupMap` or by examining the output of `dataCombined$toDataFrame()`.

More detailed information on function signatures can be found in:

- `ospsuite` documentation on:
    - [loadDataImporterConfiguration()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataImporterConfiguration.html)
    - [DataImporterConfiguration class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataImporterConfiguration.html)
    - [createImporterConfigurationForFile()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/createImporterConfigurationForFile.html)
    - [DataSet class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataSet.html)
    - [dataSetToDataFrame()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/dataSetToDataFrame.html)
    - [loadDataSetsFromExcel()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataSetsFromExcel.html)
    - [loadDataSetFromPKML()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/loadDataSetFromPKML.html)
    - [saveDataSetToPKML()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/saveDataSetToPKML.html)
    - [`DataCombined` class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataCombined.html)
    - [`plotIndividualTimeProfile()`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotIndividualTimeProfile.html)
    - [`plotObservedVsSimulated()`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotObservedVsSimulated.html)
    - [`plotResidualsVsSimulated()`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotResidualsVsSimulated.html)
    - [`plotResidualsVsTime()`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotResidualsVsTime.html)
- `tlf` documentation on:
    - [`PlotGridConfiguration`](https://www.open-systems-pharmacology.org/TLF-Library/reference/PlotGridConfiguration.html)
    - [`plotGrid()`](https://www.open-systems-pharmacology.org/TLF-Library/reference/plotGrid.html)
    - [`ExportConfiguration`](https://www.open-systems-pharmacology.org/TLF-Library/reference/ExportConfiguration.html)
    
    

<!-- ### Troubleshooting -->


<!-- More detailed information on function signatures can be found in: -->

<!-- -   `esqlabsR` documentation on: -->
<!-- -   [createDefaultProjectConfiguration()](https://esqlabs.github.io/esqlabsR/reference/createDefaultProjectConfiguration.html) -->
<!-- -   [ProjectConfiguration class](https://esqlabs.github.io/esqlabsR/reference/ProjectConfiguration.html) -->
<!-- -   [loadObservedData()](https://esqlabs.github.io/esqlabsR/reference/loadObservedData.html) -->
<!-- -   `ospsuite` documentation on: -->


