---
title: "3. Run Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{esqlabsR-run-simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)

library(esqlabsR)

projectConfiguration <- createProjectConfiguration(exampleProjectConfigurationPath())
```

Running scenarios can be divided into three steps:

1.  Creating the `ScenarioConfiguration` objects,
2.  Creating the `Scenario` objects from configurations,
3.  Running the scenarios.

### Creating `ScenarioConfiguration` objects

A `ScenarioConfiguration` object holds all information about a scenario
that is defined in the Excel files. Though it is possible to create an
empty `ScenarioConfiguration` and populate it by hand, we usually want
to create scenario configurations from the Excel files by calling the
`readScenarioConfigurationFromExcel()` function. A
`ScenarioConfiguration` is based on the `ProjectConfiguration`, which
has to be provided as an argument to the function. To create the
configuration for the *'TestScenario'* scenario defined in the
`Scenarios.xlsx` file, we call:

```{r createScenarioConfigurations}
# Create `ScenarioConfiguration` objects from excel files
scenarioConfigurations <- readScenarioConfigurationFromExcel(
  scenarioNames = "TestScenario",
  projectConfiguration = projectConfiguration
)
```

```{r, echo = FALSE}
scenarioConfigurations$TestScenario$print(projectConfiguration = FALSE)
```

Alternatively, we can create configurations for all scenarios defined in
the `Scenarios.xlsx` by calling
`readScenarioConfigurationFromExcel(projectConfiguration = projectConfiguration)`
without specifying the scenarios' names.

### Creating `Scenario` objects from configurations

Once all scenario configurations are set up, `Scenario` objects can be
created from them. A `Scenario` object contains the fully parametrized
simulation, the `Population` object in case of a population simulation,
the underlying `ScenarioConfiguration` object, and the list of all
user-defined parameters.

During the model development/fitting phase, you might want to test
parameter values before storing them in the `Parameters` Excel files.
You can define the paths of the test parameters, their values, and the
units in the `TestParameters` file and pass them as `customParams`
argument to the `createScenarios()` function.

```{r createScenarios}
# Create `Scenario` objects from `ScenarioConfiguration` objects
scenarios <- createScenarios(scenarioConfigurations)
```

You can view the final parametrization that is applied to the simulation
by calling the `finalCustomParams` property:

```{r}
scenarios$TestScenario$finalCustomParams
```

### Running scenarios

Once the `Scenario` objects are created, they can be simulated by
calling the `runScenarios()` function. The output of this function is a
named list, where the names are scenario names, and the values are the
lists of simulations, `SimulatioResults` produced by running the
simulation, the output values of the `SimulationResults`, and the
population if the scenario is a population simulation.

```{r runScenarios, results='hold'}
simulatedScenariosResults <- runScenarios(
  scenarios = scenarios
)

# Each simulation is stored separately inside the simulatedScenariosResults
names(simulatedScenariosResults)

# Each simulation can be accessed using its name
simulatedScenariosResults$TestScenario$simulation

# Of course, it contains simulated results as dataframe
head(simulatedScenariosResults$TestScenario$outputValues$data)

# It also contains dataframe's metadata
head(simulatedScenariosResults$TestScenario$outputValues$metaData)
```

It is a good idea to store simulation results as `*.csv` along with
simulations as `.*pkml` and, optionally, the applied population as
`*.csv` file and load them for further processing to avoid re-simulating
every time, e.g., a change to a figure showing the simulation results is
required. The convenient function for saving the results of a scenario
run is `saveScenarioResults`. Then, you will be able to restore the
results using `loadScenarioResults`.

Now that the results have been generated, you can proceed to the next
step in the `vignette("esqlabsR-plot-results")`.
<!-- Read the following sections to learn more about the options for running simulations. -->
