---
title: "Get Started"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to esqlabsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

## Introduction

`{esqlabsR}` is designed to make life easier for OSPS users. The workflows
implemented in this package rely on simple concepts:

  - a predefined project folder structure,
  - a JSON configuration file for definition of simulation scenarios and outputs.

## Initialize Project

`{esqlabsR}` facilitates modeling and simulation workflows with OSPS relying on
a well defined project structure. The project is configured through a
`ProjectConfiguration.json` file and supporting Excel files for detailed
parametrization (model parameters, individuals, populations, etc.).

A new project folder can be initialized with the `initProject()`
function.

```{r, eval=FALSE}
library(esqlabsR)

initProject()
```

It will create the expected simulation project's folder structure, which is described
in details in `vignette("project-structure")`.

The project is configured through a `ProjectConfiguration.json` file
located in the root of the project folder. This file defines *where* all the
necessary files are stored, which scenarios to run, and how observed data is
loaded. Supporting Excel files (e.g., `ModelParameters.xlsx`,
`Individuals.xlsx`) provide detailed parametrization for scenarios.

Loading the project configuration is the first step in any workflow using
`{esqlabsR}`.

```{r, eval=FALSE}
my_project_configuration <- loadProject(path = "path/to/ProjectConfiguration.json")
```

Note that the `ProjectConfiguration` stores paths to other files that
are used in the project.

To create an example `ProjectConfiguration` and execute the rest of this tutorial,
run the following:

```{r, warning=FALSE, message = FALSE}
projectConfigPath <- system.file(
  "extdata", "projects", "Example", "ProjectConfiguration.json",
  package = "esqlabsR"
)
my_project_configuration <- loadProject(projectConfigPath)
```

## Version Control and Project Sharing

Since `ProjectConfiguration.json` is a text-based format, it works naturally
with version control systems like Git. You can track changes, share projects
with team members, and ensure reproducibility across environments.

If you have an existing Excel-based project, you can migrate it to JSON:

```{r, eval=FALSE}
# Migrate Excel project to JSON
importProjectConfigurationFromExcel("ProjectConfiguration.xlsx")

# Export back to Excel if needed
exportProjectConfigurationToExcel(my_project_configuration)
```

For comprehensive information about version control features, workflows, and best practices, see the "Version Control and Project Sharing" section in `vignette("project-structure")`.

## Design Scenarios

Now that the project's file structure is ready, the next step is to run
simulations. In `{esqlabsR}`, simulations are run by defining and
executing multiple **scenarios**.

Scenarios are defined in the `scenarios` section of the
`ProjectConfiguration.json` file. Each scenario entry specifies:

-   `modelFile`: the name of the `.pkml` file that can be found in the
    `Models/` directory.
-   `parameterSheets`: the names of the *sheets* in
    `ModelParameters.xlsx` that contain the parameters to apply to the model,
-   `applicationProtocol`: the name of the *sheet* in
    `Applications.xlsx` that specifies the simulated
    administration protocol,
-   `simulationTime` and `simulationTimeUnit`: define the time range to
    simulate,
-   `individualId`: the id of the individual to simulate. The id and
    settings for individuals are defined in `Individuals.xlsx`,
-   `populationId`: the id of the *population* to simulate. The id and
    settings for populations are defined in `Populations.xlsx`.

Detailed parametrization (model parameters, individuals, populations,
application protocols) is still stored in Excel files referenced by the
JSON configuration.

For more information on all the scenario parameters that can be set up,
read `vignette("design-scenarios")`.

## Run Scenarios

Once all the scenario properties are set up, we run the simulations by
passing the `ProjectConfiguration` object to `runScenarios()`. We can
optionally specify which scenarios to run:

```{r}
myScenarioResults <- runScenarios(
  my_project_configuration,
  scenarioNames = "Aciclovir_iv"
)
```

Simulation results can be saved for later use: `exportScenarioResults()`
and `importScenarioResults()`.

```{r, eval = FALSE}
exportScenarioResults(myScenarioResults,
  projectConfiguration = my_project_configuration,
  outputFolder = my_project_configuration$outputFolder
)

myScenarioResults <- importScenarioResults(
  scenarioNames = "Aciclovir_iv",
  resultsFolder = my_project_configuration$outputFolder
)
```

Learn more on how to run simulations in
`vignette("run-simulations")`.

## Plot Results

The simulation results can now be plotted. For this, the package relies
on on the concept of combining multiple simulation results and observed
data using the `DataCombined` class and creating figures using functions
implemented in the `{ospsuite}` package.

First, a `DataCombined` object is initialized:

```{r}
my_datacombined <- DataCombined$new()
```

Then, simulation result are added to the dataCombined object:

```{r}
my_datacombined$addSimulationResults(myScenarioResults$Aciclovir_iv$results,
  names = "Simulated",
  groups = "Aciclovir"
)
```

Finally, the plot is generated:

```{r, out.width="80%", fig.align="center", fig.width=5, fig.asp=0.618, dpi=300}
plotIndividualTimeProfile(my_datacombined)
```

In some cases, we also want to plot the observed experimental data.
Observed data is declared in the `observedData` section of the
`ProjectConfiguration.json` file and loaded automatically when the
project is loaded. The data is available via the
`ProjectConfiguration$observedData` field.

```{r, out.width="80%", fig.align="center", fig.width=5, fig.asp=0.618, dpi=300}
observed_data <- my_project_configuration$observedData

my_datacombined$addDataSets(observed_data, names = "Observed", groups = "Aciclovir")

plotObservedVsSimulated(my_datacombined)
```

Many other plot types are available in `{esqlabsR}`, read
`vignette("plot-results")` to learn more.
