---
title: "Running simulations"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Example workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  eval = FALSE
)
```

```{r, echo = FALSE, results = "hide", message = F}
library(esqlabsR)
```

Within the `esqlabsR` framework, the simulations are run by defining and executing multiple *scenarios*. To define a scenario, create an Excel file located in `projectConfiguration$scenarioDefinitionFile` that lists the scenario name, involved individuals, application protocols, simulation time, PKML model and model parameters. An example of the scenario definition used in this example is available [on github](https://github.com/esqLABS/esqlabsR/blob/HEAD/tests/data/TestProject/Parameters/Scenarios.xlsx).

Scenarios are then executed by creating a `ScenarioConfiguration` class instance and calling a `runScenarios()` function:

```{r}
scenarioConfiguration <- ScenarioConfiguration$new(projectConfiguration)
OutputPaths <- enum(list(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
))
simulatedScenarios <- runScenarios(scenarioNames = c("TestScenario"),
                                   scenarioConfiguration = scenarioConfiguration, 
                                   customParams = NULL, saveSimulationsToPKML = FALSE)
```

The `runScenarios()` function relies on an `OutputPaths` variable being defined in the current environment. The variable is expected to contain [references to an entity](https://docs.open-systems-pharmacology.org/working-with-mobi/mobi-documentation/model-building-components) that will be used as an output variable for the simulation.

The `runScenarios()` function returns a named list of lists with `Simulation` class instances, `SimulationResults` class instances, and vectors of output values.

More detailed information on function signatures can be found in:

* `esqlabsR` documentation on:
    * [ProjectConfiguration class](https://esqlabs.github.io/esqlabsR/reference/ProjectConfiguration.html)
    * [ScenarioConfiguration class](https://esqlabs.github.io/esqlabsR/reference/ScenarioConfiguration.html)
    * [readScenarioConfigurationFromExcel()](https://esqlabs.github.io/esqlabsR/reference/readScenarioConfigurationFromExcel.html)
    * [runScenarios()](https://esqlabs.github.io/esqlabsR/reference/runScenarios.html)
    
* `ospsuite` documentation on:
    * [Simulation class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/Simulation.html)
    * [SimulationResults class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/SimulationResults.html)
