---
title: "Visualization and reporting"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Visualization and reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = "../tests/data/TestProject/Code/"
)
```

```{r, echo = FALSE, results = "hide", message = F}
library(esqlabsR)
projectConfiguration <- createDefaultProjectConfiguration()
dataSheets <- "Laskin 1982.Group A"
observedData <- loadObservedData(projectConfiguration = projectConfiguration, sheets = dataSheets)
scenarioConfiguration <- ScenarioConfiguration$new(projectConfiguration)
OutputPaths <- enum(list(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
))
simulatedScenarios <- runScenarios(
  scenarioNames = c("TestScenario"),
  scenarioConfiguration = scenarioConfiguration,
  customParams = NULL, saveSimulationsToPKML = FALSE
)
```

### Workflow

Plotting the simulation results is an important part of model diagnostics and quality control. Simulated modeling scenarios can be passed to plotting functions from the `{ospsuite}` package to create uniformly-looking plots.

`DataCombined` is a class used to store matching observed and simulated data. Initialize a new class instance and populate it with data with the following code:

```{r datacombined}
dataCombined <- DataCombined$new()
dataCombined$addDataSets(observedData, names = "Observed", groups = "Aciclovir")
dataCombined$addSimulationResults(simulatedScenarios$TestScenario$results, names = "Simulated", groups = "Aciclovir")
dataCombined$toDataFrame()
```

The simulation results are stored in a list returned by the `runScenarios()` function. Plotting and visualization is performed by storing these results along with matching observed data in an instance of the `DataCombined` class and passing it to plotting functions.

Plotting functions in the `{ospsuite}` package are wrappers around `{tlf}` plotting functions that provide default plot configuration options. All of them accept instances of `DataCombined` class as the data source. 

Time profile plots visualize the pharmacokinetics of the drug in question and help assess if the observed data (shown with scatter plots) match the simulated data (shown with line plots).

```{r plot time profile, fig.width=7, fig.height=4}
plotIndividualTimeProfile(dataCombined)
```

Observed vs simulated plots show if simulated time points and observed time points follow a linear trend.

```{r plot obs vs pred, fig.width=7, fig.height=4}
plotObservedVsSimulated(dataCombined)
```

Residual plots show if there is a systematic bias in how the simulation represents values either in high-concentration or low-concentration regions, or, alternatively, in early or late time periods.

```{r residuals vs simulated, fig.width=7, fig.height=4}
plotResidualsVsSimulated(dataCombined)
```

```{r residuals vs time, fig.width=7, fig.height=4}
plotResidualsVsTime(dataCombined)
```

The resulting plots (returned by plotting functions) are instances of `ggplot` class. They can be modified and saved to files with `{ggplot2}` functions:

```{r save plots}
plotObject <- plotIndividualTimeProfile(dataCombined)
ggplot2::ggsave(filename = "../Results/aciclovir_time_profile.png", plotObject, width = 8, height = 4)
```

### Troubleshooting

* At any time, you can check the groups assigned to the datasets in the `DataCombined` class instance by examining the output of `dataCombined$toDataFrame()`.

More detailed information on function signatures can be found in:

* `ospsuite` documentation on:
    * [Simulation class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/Simulation.html)
    * [SimulationResults class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/SimulationResults.html)
    * [DataCombined class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataCombined.html)
    * [plotIndividualTimeProfile()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotIndividualTimeProfile.html)
    * [plotObservedVsSimulated()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotObservedVsSimulated.html)
    * [plotResidualsVsSimulated()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotResidualsVsSimulated.html)
    * [plotResidualsVsTime()](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/plotResidualsVsTime.html)
