---
title: "Creating figures"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Creating figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE, results = "hide", message = F}
library(esqlabsR)
```

This package offers a workflow for creating standardized figures for time-values data, such as time-concentration profiles, goodness-of-fit plots, etc. The integral part of these workflows are the objects of the class `DataMapping`. A `DataMapping` organizes the objects of the class `XYData` (see [Data handling](data-handling.html) for more information) into groups and maps simulation results to observed data.

## `DataMapping` overview
The first step in organizing the data in a `DataMapping` object and creating figures is the creation of an empty `DataMapping` object.

```{r createDataMapping}
dataMapping <- DataMapping$new()
print(dataMapping)
```

In the next step, the mapping is populated with xy-values series that can be organized into groups.
