---
title: "How to run a sensitivity analysis"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How to run a sensitivity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.showtext = TRUE
)
library(esqlabsR)
```

Sensitivity analysis varies model parameters and measures the effect on PK outcomes, which helps you judge model robustness when input values are uncertain. This vignette walks through a complete workflow with `sensitivityCalculation()` and its plot functions, using **aciclovir** as the example: run a base simulation, vary selected parameters, calculate sensitivities, and visualize the impact. For the mathematical background, see the OSPS documentation on [sensitivity analysis](https://docs.open-systems-pharmacology.org/shared-tools-and-example-workflows/sensitivity-analysis).

## Running the sensitivity analysis

`sensitivityCalculation()` operates on an OSPS `Simulation` object, not on a project. You obtain one by loading the worked example project and running its `aciclovir_iv` scenario; the resulting Scenario Result carries the prepared `Simulation` in its `simulation` field:

```{r load}
project <- loadProject(exampleProjectPath())
results <- runScenarios(project, scenarios = "aciclovir_iv")
simulation <- results[["aciclovir_iv"]]$simulation
```

By default `sensitivityCalculation()` re-runs the simulation with each parameter scaled from 0.1x to 10x its original value; pass `variationRange` to change this. It returns the varied parameters, the simulation results, and a data frame of PK metrics per variation.

Here you analyze how changes in *lipophilicity*, *dose*, and *glomerular filtration rate (GFR fraction)* affect the pharmacokinetics of aciclovir:

```{r analysis}
outputPaths <- c(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
)
parameterPaths <- c(
  "Lipophilicity" = "Aciclovir|Lipophilicity",
  "Dose" = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  "GFR fraction" = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Glomerular Filtration-GFR-Aciclovir|GFR fraction"
)

analysis <- sensitivityCalculation(simulation, outputPaths, parameterPaths)
```

To interpret the results, focus on *lipophilicity*. Isolate that parameter and the columns that drive the sensitivity calculation:

```{r}
analysis$pkData |>
  dplyr::filter(
    ParameterPath == "Aciclovir|Lipophilicity",
    PKParameter == "AUC_inf",
    ParameterFactor %in% c(0.1, 1, 10)
  ) |>
  dplyr::select(
    ParameterFactor, PKParameterValue,
    PKPercentChange, SensitivityPKParameter
  )
```

In this example the default lipophilicity (−0.1 log units) yields an AUC of 3916.4 µmol·min/L. A ten-fold higher value (−0.01) reduces AUC by 0.52 %, whereas a ten-fold lower value (−1.0) increases it by 2.55 %. The sensitivity of AUC to a 10-fold decrease in lipophilicity is calculated as:
$$ \text{Sensitivity}= \frac{\Delta \text{AUC}}{\Delta \text{lipophilicity}}\times\frac{\text{lipophilicity}}{\text{AUC}} = \frac{4016.3-3916.4}{-1.0--0.1} \times \frac{-0.1}{3916.4} = 0.0028 $$

## Saving and loading sensitivity results

Save the results of a sensitivity analysis with `saveSensitivityCalculation()` and restore them with `loadSensitivityCalculation()`. The `simulation` argument in `loadSensitivityCalculation()` is optional: if the original `Simulation` object is provided, reloading from disk is avoided. Otherwise, the function attempts to load the `.pkml` file from the path recorded during the original analysis, which must still be accessible.

```{r save-load, eval=FALSE}
# Save to disk
outputDir <- file.path(tempdir(), "sensitivity-results")
saveSensitivityCalculation(analysis, outputDir)

# Reload from disk
analysis <- loadSensitivityCalculation(outputDir, simulation)
```

## Visualizing sensitivity results

Visualize the results of the sensitivity analysis with the following functions.

### Sensitivity spider plot

`sensitivitySpiderPlot()` visualizes how PK parameters respond as model input parameters are scaled. Each panel represents one PK metric (default: `C_max`, `t_max`, `AUC_inf`), making it straightforward to identify nonlinear trends and compare the direction and magnitude of effects.

```{r spider-plot, fig.height=4}
sensitivitySpiderPlot(analysis)
```

In this example, *Dose* shows a strong and nonlinear effect on both `C_max` and `AUC_inf`, while `t_max` remains relatively stable. *GFR fraction* and *lipophilicity* have more modest but still visible effects.

### Sensitivity time profiles

`sensitivityTimeProfiles()` displays full concentration-time curves for each input parameter variation. This plot is suited to understanding how parameter changes affect the shape and timing of the drug profile, rather than summary PK metrics.

Each panel corresponds to one input parameter, with a color gradient representing the scaling factor: 0.1 = blue, 1.0 = black, 10 = red.

```{r time-profile, fig.height=4}
sensitivityTimeProfiles(analysis)
```

Here, increasing the *Dose* uniformly raises the profile, while changes in *GFR fraction* affect the rate of decline, indicating faster or slower elimination. *Lipophilicity* causes only minor shifts in the curve.

### Sensitivity tornado plot

`sensitivityTornadoPlot()` provides a compact, side-by-side comparison of how each input parameter influences different PK outputs at a fixed scaling factor. The plot is best used to **rank parameters by influence** and identify those with the strongest impact.

By default, the plot compares the results at 0.1x and 10x the original parameter value. Other scaling factors can also be used, but they must be included in the `variationRange` passed to `sensitivityCalculation()`.

```{r tornado-plot, fig.height=3.5}
sensitivityTornadoPlot(analysis)
```

The tornado plot confirms the trends seen in the previous plots, but makes it easier to compare the relative impact of each parameter across PK metrics.

To change colors, scales, titles, or which PK parameters appear in any of these plots, see `vignette("how-to-customize-sensitivity-plots")`.

## Sensitivity calculation for user-defined (non-PK) outputs

Though the typical sensitivity-analysis workflows focus on PK parameters, you can also calculate sensitivities for any numerical model output using user-defined functions. As for the PK parameters, the provided function(s) are applied to each output defined in the `outputPaths` argument of `sensitivityCalculation()`.

The custom functions are provided as a named list in the `customOutputFunctions` argument. Each function must take arguments `x` and `y`, through which it accesses the simulated time values (`x`) and the output values (`y`), and must return a single numeric value.

In the following example, you calculate the sensitivity of the average concentration of aciclovir in the peripheral venous blood to the lipophilicity of aciclovir. First, define a function that calculates the mean of a numeric vector:

```{r custom-function}
meanFunction <- function(x, y) {
  mean(y)
}
```

Next, run the sensitivity analysis for the average concentration, reusing the `simulation` from the base run. To define custom labels for the parameters in the resulting plots, pass a *named vector* to `parameterPaths`, where the names are used as labels instead of full paths:

```{r custom-analysis}
customOutputPaths <- c(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
)
customParameterPaths <- c("Lipophilicity" = "Aciclovir|Lipophilicity")

customAnalysis <- sensitivityCalculation(
  simulation,
  customOutputPaths,
  pkParameters = c("C_max"),
  parameterPaths = customParameterPaths,
  customOutputFunctions = list(AverageConcentration = meanFunction)
)
```

[**Note**]{.underline}: To calculate the sensitivity of the custom function only, without the default PK parameters, set `pkParameters` to an empty list.

Finally, plot the sensitivity of the average concentration of aciclovir in the peripheral venous blood to its lipophilicity:

```{r custom-plot, fig.height=4}
sensitivitySpiderPlot(customAnalysis)
```

## Troubleshooting

-   The `SensitivityPKParameter` column in `analysis$pkData` is `NA` in the rows corresponding to the initial parameter values (that is, where the multiplication factor is 1).

## Where to go next

To customize the appearance of these plots (colors, scales, titles, and which PK parameters appear), see `vignette("how-to-customize-sensitivity-plots")`.
