---
title: "How to validate a project"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How to validate a project}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

Before you run anything, check that the project hangs together: every reference resolves, no required field is missing, no id is duplicated. `validateProject()` does this and prints a readable report. This guide shows how to run it, read the report, act on a problem, and gate a script on the result.

It assumes you know what a Project is and how to author one (see `vignette("how-to-set-up-a-project")` and `vignette("how-to-design-scenarios")`).

Run it right after authoring and before running: `runScenarios()` and `createPlots()` each validate the parts of the project their own work is built from, so an explicit `validateProject()` is what gives you the whole picture at the moment you choose, rather than the first blocking error a slow run happens to hit. `vignette("about-the-project-model")` covers why validation is split up this way and why it reports two levels of finding.

Throughout, work on a writable copy of the bundled example:

```{r}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))
```

## Read a report

A report separates two kinds of finding, and which one you got decides whether you have to act:

- A **Critical Error** blocks. A duplicate id, a required field left empty, or a reference to something that does not exist. The section's `isValid()` returns `FALSE`, and the operations built on it will not run until you fix it.
- A **Warning** does not block. It is surfaced so you see it, and a section with warnings and no critical errors is still valid.

The quickest single question, "is anything blocking?", is `isAnyCriticalErrors()`.

Passing `validate = FALSE` to `runScenarios()` or `createPlots()` skips their check and lets a scenario build on whatever the model file already carries. Reach for it only when you know the gap is deliberate.

## Validate a clean project

The bundled example is well formed, so its report is short. `validateProject()` runs every section validator plus a cross-reference pass and returns a structured object that prints a summary:

```{r}
results <- validateProject(project)
results
```

`isAnyCriticalErrors()` agrees:

```{r}
isAnyCriticalErrors(results)
```

## Diagnose a dangling reference

The most common authoring slip is a dangling reference: a definition pointing at an id that does not exist, usually a small misspelling. Note that a *loaded* project can hold one even though `addScenario()` would have refused it, because reading a project reports gaps rather than rejecting them (`vignette("about-the-project-model")` explains why).

To see what the report looks like, suppose one scenario file in our copy mistypes the individual id `adult_male` as `adlt_male`. Reloading the project catches the dangling reference at load time, and `validateProject()` then reports it in full:

```{r, echo = FALSE}
scenario_file <- file.path(
  project_dir, "definitions", "scenarios", "aciclovir_iv.json"
)
text <- readLines(scenario_file)
text <- gsub('"adult_male"', '"adlt_male"', text, fixed = TRUE)
writeLines(text, scenario_file)

broken <- suppressWarnings(loadProject(file.path(project_dir, "Project.json")))
```

The report is grouped by definition type: a cross (`✖`) marks each critical error, an exclamation mark (`!`) marks each warning, and the sections with nothing to report collapse into a single tail line.

```{r}
broken_results <- validateProject(broken)
broken_results
```

Notice the message does not just say the reference is missing; it offers a **"did you mean"** suggestion built from the ids that actually exist (`did you mean 'adult_male'?`), the validator catching the most common mistake, a small misspelling of an existing id.

## Read the result in code

The printed report is for you; the object `validateProject()` returns is for your code. It is a named list with one entry per definition type, and each entry is a Validation Result you can index into. So a script can gate on a specific section without parsing any text:

```{r}
names(broken_results)
broken_results$crossReferences$critical_errors[[1]]$message
```

`validationSummary()` rolls the per-section results into aggregate counts and the names of the sections that produced errors or warnings, which is handy for a one-line status in a report or a log:

```{r}
validationSummary(broken_results)
```

## Resolve and re-validate

Fixing the problem is the same authoring you already know. Here the scenario simply needs to point at the correct individual id; `setScenario()` updates it in memory (call `saveProject()` when you want the fix on disk):

```{r}
setScenario(broken, "aciclovir_iv", individual = "adult_male")
```

Re-validate to confirm the project is clean again:

```{r}
fixed_results <- validateProject(broken)
fixed_results
isAnyCriticalErrors(fixed_results)
```

Both the report and `isAnyCriticalErrors()` are clean, so the project is ready to run.

## Where to go next

With a clean project in hand, see `vignette("how-to-run-simulations")` to execute the scenarios and `vignette("how-to-plot-results")` to turn the results into figures. To author or repair the definitions a validation problem points at, see `vignette("how-to-design-scenarios")`.

To look up which fields a definition requires, see `vignette("reference-definition-fields")`. For why validation reports two levels and checks a different slice of the project per entry point, see `vignette("about-the-project-model")`.
