---
title: "How to design scenarios"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How to design scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

A **Scenario** is the unit of work in `esqlabsR` - a well-defined simulation setup. A scenario defines the model (a `.pkml` simulation file exported from PK-Sim or MoBi), individual characteristics, a population (optionally), the administration protocol(s), the parameters to apply, the outputs to record, and the time grid. A **Scenario** is given by a set of **definitions**. You can define scenarios (and the definitions they reference) in a **Project**, and pass the Project to `runScenarios()` to execute them, or to other workflows that use the scenarios. This article shows how to build scenarios and their supporting definitions from R, and how to edit, rename, duplicate, and remove them within a Project context.

This article assumes you know what a **Project** is and how to load one; see `vignette("how-to-set-up-a-project")` if not.

## Different ways to define and edit scenarios

This article covers the functions and methods that allow to create and edit scenarios (and the underlying definitions) using R code. Other options include editing the Excel files, or using dedicated applications like the ESQapp (available in ESQlabs subscription).

## Following the examples in this article

Every `add*`, `set*`, and `remove*` function in this article changes the project in your R session, and `saveProject()` writes those changes to the files in the project's `definitions/` folder. If you want to follow the examples, create a fresh copy of the example project provided with the package in a temporary directory and load that:

```{r}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))
project
```

In your own work you would point `loadProject()` at your real project instead; everything below operates on this `project` object. Each kind of definition has its own section under `project$definitions`. Print a section to see the ids it holds, and pick out one definition by its id, as in `project$definitions$scenarios[["aciclovir_iv"]]`:

```{r}
project$definitions$scenarios
```

## The simplest scenario

The minimal amount of information a scenario definition needs is the model to run. Without other fields specified, it uses what PK-Sim or MoBi exported: the dosing written into the model, the model's own individual, its parameter values, the outputs it records, and its time grid. Only the solver settings can come from elsewhere: if the project sets `defaultSolverSettings` (for example `project$defaultSolverSettings <- list(relTol = 1e-6)`; see `vignette("reference-project-file")`), those apply. The example project sets none, so such a scenario simulates the model exactly as it was exported:

```{r}
# Create a new scenario for the project
addScenario(project, "aciclovir_as_exported", modelFile = "Aciclovir.pkml")
project$definitions$scenarios[["aciclovir_as_exported"]]
```

`modelFile` names the exported simulation file in the project's simulations folder (`Models/Simulations` in the example project). This one field is enough to run the scenario: `runScenarios(project, scenarios = "aciclovir_as_exported")`, covered in `vignette("how-to-run-simulations")`.

Printing a scenario shows its settings. `Name` is the scenario's id, `Type` says whether it simulates an individual or a population, and `Administration Protocol` is its application. A field shown as `<empty string>` (or  `<empty vector>`) is not set, so what the model file carries is used instead. The time grid can be printed with `project$definitions$scenarios[["aciclovir_as_exported"]]$simulationTime`: `NULL` means the model file's own time grid is used.

```{r}
# Print scenario information
print(project$definitions$scenarios[["aciclovir_as_exported"]])

# Print the time grid (if defined separately)
print(project$definitions$scenarios[["aciclovir_as_exported"]])$simulationTime
```

The following sections of this article describe how to set up scenarios that are based on the exported model, but differ in their properties.

## Building the supporting definitions

A scenario can only reference definitions (e.g., an individual) that already exist in the project, so before you create the scenario you typically define the pieces it will use.

When you assign a new id (i.e., its unique name) to a definition, it is made lowercase and safe as a file name, because the id is also the name of its file. The function warns you when that changes what you typed, as the `addOutputPath()` call below shows. `vignette("about-the-project-model")` explains why.

### Output paths

An **Output Path** definition maps a short id to a full model path in OSPS notation. Recording results by id keeps scenario definitions readable and lets several scenarios share the same output:

```{r, warning = TRUE}
addOutputPath(
  project,
  id = "Aciclovir/Liver:Conc",
  path = "Organism|Liver|Periportal|Intracellular|Aciclovir|Concentration in container"
)
project$definitions$outputPaths
project$definitions$outputPaths[["aciclovir_liver_conc"]]
```

The id you typed was stored as `aciclovir_liver_conc`. An id that is already lowercase and safe as a file name is stored as typed, without a warning:

```{r}
addOutputPath(
  project,
  id = "aciclovir_kidney",
  path = "Organism|Kidney|Intracellular|Aciclovir|Concentration in container"
)
```

**Note:** So far, we created a new definition of an output path, but we did not specify this output in any of the scenarios. This definition can now be used across the project, e.g., for defining outputs that are generated in a scenario, mapping of observed data to simulation results in plots definitions, or defining which model outputs to use in a parameter identification.

### Parameter sets

Model parameters are organized into named **Parameter Sets**. A parameter set is a group of parameter entries that are applied to a simulation. Multiple parameter sets can be combined in one scenario.

You create an empty set with `addParameterSet()` and fill it with `addParameterEntry()`. Each entry names the container path, the parameter name, the value, and its unit. To add several entries in one call, give each of these as a vector with one element per entry. All the vectors must have the same length:

```{r}
addParameterSet(project, "ckd")
addParameterEntry(
  project,
  "ckd",
  containerPath = "Organism|Kidney",
  parameterName = "GFR",
  value = 30,
  units = "ml/min"
)
project$definitions$parameterSets[["ckd"]]
```

Scenarios, individuals, and applications all take their parameter sets from this one collection, so a set such as `global` can be shared by many scenarios. `addParameterEntry()` creates the set if it does not exist yet, and stops with an error if the set already has an entry for the same parameter (the same `containerPath` and `parameterName`), unless you pass `overwrite = TRUE`.

To change a value in an existing set, call `addParameterEntry()` again with `overwrite = TRUE`. To take an entry out of a set, use `removeParameterEntry()`. If that was the last entry, the set itself is removed too, and anything that still names the set is left with a broken reference (you get a warning naming them). `removeParameterSet()` removes a whole set from the project.

### Initial conditions

An **Initial Condition** set gives molecules their amount at the start of the simulation. It works like a parameter set: create an empty set with `addInitialConditions()`, then add entries with `addInitialConditionEntry()`, which, like `addParameterEntry()`, takes vectors to add several entries in one call. Every entry needs a path, a value, and a unit:

```{r}
addInitialConditions(project, "plasma_start")
addInitialConditionEntry(
  project,
  "plasma_start",
  path = "Organism|VenousBlood|Plasma|Aciclovir",
  value = 0.5,
  unit = "µmol"
)
project$definitions$initialConditions[["plasma_start"]]
```

A scenario uses the set through its `initialConditions` argument.

### Individuals

An **Individual** carries the biometrics of a single simulated subject plus any individual-specific parameter sets. When an individual is defined in a scenario, a parameter set describing the physiology of the defined individual is applied to the simulation. This way it is possible to simulate different individuals (including different populations) using the same original simulation. An individual can be of a different species than the model. A scenario that uses a Beagle, Dog, Minipig, Monkey, Mouse, Rabbit, or Rat individual with a model exported for a human scales the model to that species: it takes the species' values from PK-Sim (organ volumes, blood flows, tissue composition, gut physiology, lumen pH), and these follow the individual's body weight. Scaling between two animal species, for example from a rat model to a mouse, works the same way. A model exported for an animal cannot be scaled to a human: the scenario stops with an error when it is run.

For the definition of an individual, Only `species` is required; every other field is optional, and an omitted biometric is taken from the PK-Sim defaults for that species (`vignette("reference-definition-fields")` lists the defaults). Here we add a patient with chronic kidney disease:

```{r}
addIndividual(
  project,
  "adult_male_ckd",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "MALE",
  weight = 73,
  height = 176,
  age = 60,
  proteinOntogenies = "CYP3A4:CYP3A4",
  parameterSets = c("adult_male_default", "ckd")
)
project$definitions$individuals[["adult_male_ckd"]]
project$definitions$individuals[["adult_male_ckd"]]$proteinOntogenies
```

The `parameterSets` argument is a character vector of ids referencing the project's parameter sets; it is validated immediately, so a typo in a set id is caught. The sets are applied in the order you list them, so `ckd` comes after `adult_male_default`, and its GFR of 30 ml/min replaces the 90 ml/min that `adult_male_default` sets.

`proteinOntogenies` sets how the amount of an enzyme or transporter changes with age. Each entry pairs a protein in the model with one of PK-Sim's standard ontogenies, written `"Protein:Ontogeny"`; `ospsuite::StandardOntogeny` lists them. A protein you do not list keeps the ontogeny stored in the model file. Above, the model's CYP3A4 gets PK-Sim's CYP3A4 ontogeny, as in the example project's `adult_male`. Here that changes nothing: the Aciclovir model already carries this ontogeny, and a 60-year-old has the full adult amount. You need the field when a protein in the model has no ontogeny, or when you want a different one, and it matters most for children.

To add several individuals in one call, pass a vector of ids. The id sets the count, and the other arguments are recycled or aligned against it; `vignette("reference-definition-fields")` gives the exact rules. If any one definition fails validation, none is applied.

```{r}
addIndividual(
  project,
  c("teen_male", "teen_female"),
  species = "Human",
  population = "European_ICRP_2002",
  gender = c("MALE", "FEMALE"),
  weight = c(50, 48),
  height = 160,
  age = 14
)
names(project$definitions$individuals)
```

### Populations

A **Population** describes a virtual population to sample: a species, a number of individuals, and optional demographic ranges (age, weight, height, BMI) and their units. A scenario that references a population becomes a population simulation:

```{r}
addPopulation(
  project,
  "pediatric",
  species = "Human",
  population = "European_ICRP_2002",
  numberOfIndividuals = 25,
  ageMin = 2,
  ageMax = 12,
  proportionOfFemales = 50
)
project$definitions$populations[["pediatric"]]
```

Like an individual, a population takes `proteinOntogenies`; each virtual individual then gets the ontogeny factor for its own age.

### Applications

An **Application** names the parameter sets that change the dosing in the model file, such as the dose, the start time, or the infusion time. It can only change administrations the model already contains: a parameter path that is not in the model stops the run. Here a new parameter set doubles the dose of the model's IV infusion. In the model, the event is called `IV 250mg 10min`.

```{r}
addParameterSet(project, "aciclovir_iv_500mg")
addParameterEntry(
  project,
  "aciclovir_iv_500mg",
  containerPath = "Events|IV 250mg 10min|No formulation|Application_1|ProtocolSchemaItem",
  parameterName = "Dose",
  value = 500,
  units = "mg"
)
addApplication(
  project,
  "aciclovir_iv_500mg",
  parameterSets = "aciclovir_iv_500mg"
)
project$definitions$applications[["aciclovir_iv_500mg"]]
```

## Creating a scenario

With the supporting definitions in place, `addScenario()` assembles them into a runnable scenario.

```{r}
addScenario(
  project,
  "aciclovir_iv_ckd",
  modelFile = "Aciclovir.pkml",
  individual = "adult_male_ckd",
  application = "aciclovir_iv_250mg",
  parameterSets = c("global", "aciclovir"),
  outputPaths = c("aciclovir_pvb", "aciclovir_kidney"),
  simulationTime = "0, 24, 60",
  simulationTimeUnit = "h"
)
project$definitions$scenarios[["aciclovir_iv_ckd"]]
```

The id is also the scenario's name. Choose an id you can read later. An id such as `Aciclovir_IV-2.5mg/kg` is stored as `aciclovir_iv-2.5mg_kg`, and the original spelling is not kept.

As in the simplest scenario above, only `modelFile` is required. Each reference or setting you leave out will use the defaults defined in the simulation, except two: steady state is off unless you turn it on, and the solver settings come from the project's `defaultSolverSettings` where it has any. `vignette("reference-definition-fields")` lists every field with what happens when you omit it.

Three behaviors are worth knowing while you write the call:

- **References are checked as the scenario is added.** If `individual`, `population`, `application`, or any entry in `parameterSets`, `initialConditions`, or `outputPaths` names a definition that does not exist, `addScenario()` stops with an error, suggests the closest existing id, and creates nothing. This is why you build the supporting definitions first. The model file is not checked at this point: `validateProject()` warns if it is missing.
- **Parameter sets are applied in layers.** First the scenario's own `parameterSets`, in the order you list them, then the parameter sets of its individual, then those of its application. Where two sets change the same parameter, the one applied last wins. Above, `global` and `aciclovir` go on first, then the individual's `adult_male_default` and `ckd`, then the application's `aciclovir_iv_250mg_default`. That is why `ckd` belongs on the individual: listed in the scenario's `parameterSets`, its GFR would be replaced by the one in `adult_male_default`.
- **`simulationTime` is an interval string**, `"start, end, resolution"`, where resolution is points per unit of `simulationTimeUnit`. Separate several intervals with `;` to simulate part of the time course densely and the rest coarsely, as in `"0, 20, 60; 20, 504, 1"`.

Like every other edit, the new scenario exists only in your R session until you call `saveProject()`, which writes it to `definitions/scenarios/aciclovir_iv_ckd.json`.

### Population scenarios

To run a population scenario, pass a `population` id during scenario creation, or add a population later to the field `project$definitions$scenarios[["scenario_name"]]$populationId`; the scenario type changes automatically. If you also keep an `individual`, its parameter sets still apply.

`readPopulationFromCSV = TRUE` makes a population scenario read its population from a CSV file in the project's populations folder instead of sampling it. The SCV file must be named as the `populationID` defined in the scenario.

### Simulating a steady-state

If `steadyState = TRUE`, the simulation will be run for `steadyStateTime` after its final parametrization, the last simulated values of all molecules and state-variable parameters will be applied as their initial values, and the simulation will be run again. This is particularly useful when applying parametrization that changes the steady-state of the system, e.g., production rates of endogenous substrates.

### Solver properties (scenario-specific)

To change how the solver behaves for this scenario only, pass `solverSettings`, for example `solverSettings = list(relTol = 1e-6)`. See `vignette("reference-definition-fields")` for supported properties.

## Changing an existing scenario

You can update an existing scenario using the function `setScenario()`. It takes the same arguments as `addScenario()` except `overwrite`, and differs in two ways:

- **Only the fields you pass change.** Every argument you omit keeps its current value (a partial update).
- **Passing `NULL` clears an optional field**. (The required `modelFile` cannot be cleared.)

In the next example, we change the administration protocol and widen the time grid, leaving everything else as it was. Keep in mind that the administration protocol `aciclovir_iv_500mg` already exists in the project.

```{r}
setScenario(
  project,
  "aciclovir_iv_ckd",
  application = "aciclovir_iv_500mg",
  simulationTime = "0, 48, 60"
)
project$definitions$scenarios[["aciclovir_iv_ckd"]]$simulationTime
```

In the next example, we remove the individual assignment from the scenario by passing `NULL`. The consequence is that parameters describing the physiology will not be overwritten, and individul-specific parameter sets defined in `project$definitions$individuals$adult_male_ckd$parameterSets` will not be applied.

 In R, a scenario record calls this field `individualId`, not `individual`; `vignette("reference-definition-fields")` lists the fields whose names differ:

```{r}
setScenario(project, "aciclovir_iv_ckd", individual = NULL)
project$definitions$scenarios[["aciclovir_iv_ckd"]]$individualId
```

The same update logic applies to the supporting definitions.

```{r}
setIndividual(project, "adult_male_ckd", weight = 70, age = 62)
project$definitions$individuals[["adult_male_ckd"]]

setOutputPath(
  project,
  "aciclovir_kidney",
  path = "Organism|Kidney|Urine|Aciclovir|Concentration"
)
project$definitions$outputPaths[["aciclovir_kidney"]]
```

## Renaming, duplicating, and removing scenarios

A common move is to take a working scenario as a template for a new scenario. `duplicateScenario()` makes an independent copy under a new id:

```{r}
duplicateScenario(project, id = "aciclovir_iv", newId = "aciclovir_iv_highdose")
names(project$definitions$scenarios)
```

`renameScenario()` gives a scenario a new id:

```{r}
renameScenario(project, "aciclovir_iv_highdose", "aciclovir_iv_high")
names(project$definitions$scenarios)
```

Use `removeScenario()` to remove an existing scenario from the project.

```{r}
removeScenario(project, "aciclovir_iv_high")
names(project$definitions$scenarios)
```

Renaming or removing a scenario does not update the DataCombined entries or parameter-identification tasks that name it. You get a warning, and `validateProject()` reports them as critical errors until you update them.

## Creating scenarios from model files

An easy way to create a scenario for a simulation from a `.pkml` file is given by the function `createScenariosFromPKML()`. It reads each model file, extracts the output selections and the simulation time grid embedded in it, and adds a scenario to the project. Optionally, any property of a scenario can also be defined in the same call, including overwriting the output paths or the time grid.

```{r}
pkml_path <- file.path(project_dir, "Models", "Simulations", "Aciclovir.pkml")
createScenariosFromPKML(
  pkmlFilePaths = pkml_path,
  project = project,
  scenarios = "aciclovir_from_pkml"
)
project$definitions$scenarios[["aciclovir_from_pkml"]]
```

For each model, the call:

- Registers the output paths found in the model, reusing an existing id when the literal path is already known and generating a readable id otherwise.
- Makes the new scenario names unique against the scenarios already present.
- Applies the provided scenario settings (individual, population, application, parameter sets).

Pass a vector of `pkmlFilePaths` to create several scenarios at once. Pass `scenarios` to name them yourself; otherwise each scenario is named after the simulation in its model file, made lowercase and safe as a file name: the simulation `Vergin 1995 IV` becomes the scenario `vergin_1995_iv`.

## Saving your work

Every edit above exists only in your R session until you save it. Call `saveProject(project)` to write the edits to the `definitions/` folder, or `reloadProject(project)` to discard them and read the project files again. Edits you have not saved are lost when the R session ends; printing the project shows `[unsaved changes]` while there are any.

## Where to go next

Next, validate what you built with `vignette("how-to-validate-a-project")`, then run it with `vignette("how-to-run-simulations")` and plot it with `vignette("how-to-plot-results")`.

For the fields of each definition kind, their defaults, see `vignette("reference-definition-fields")`.
