---
title: "How to use observed data"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How to use observed data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

Comparing a simulation against measured data is what turns a model run into a diagnostic. In `esqlabsR`, measured data is **Observed Data**: a named series of time/value points, the package's term for what OSPS implements as a `DataSet`. This guide shows how to declare each of the four kinds of source, load the data, and remove a source.

It assumes you already know what a Project is and how to load one (see `vignette("how-to-set-up-a-project")`). To pair observed data with simulated curves in a figure, see `vignette("how-to-plot-results")`.

A project stores a **source definition**, not the numbers: a small record under `definitions/observed-data/<id>.json` saying where the data comes from and how to read it. The data itself is loaded on demand, so it always reflects the current source file. `vignette("about-the-plotting-model")` explains why, and `vignette("reference-definition-fields")` lists the required fields per source type.

The four types differ only in where the data comes from: `excel` reads sheets of a workbook through an OSPS data-importer configuration, `pkml` reads a saved `DataSet`, `script` runs an R script that returns one, and `programmatic` takes a `DataSet` you built in the session.

Throughout, scaffold a writable copy of the bundled example so the sources you add are written to a project you own, not the installed package:

```{r}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))
```

The example already ships with one observed-data source, an Excel source. You can see it as a definition on the project:

```{r}
project$definitions$observedData[[1]]
```

## Choose the handle a source is known by

Two strings attach to every source, and they are not interchangeable. The **source id** is what `removeObservedData()` matches and what names the definition file. The **DataSet name** is what a DataCombined references and what `getObservedDataNames()` lists; it comes from the data itself, so an Excel sheet becomes a long, fully qualified name. `vignette("about-the-plotting-model")` explains why one string cannot serve both roles.

The working rule: **read the DataSet name from `getObservedDataNames()` rather than typing it**, and use the source id only to remove a source.

Set an `id` when you want to choose that handle yourself instead of inheriting it from the file. Since the id also names the definition file, the declaration below is stored as `definitions/observed-data/laskin_iv.json` rather than under the data file's name:

```{r, eval = FALSE}
addObservedData(
  project,
  list(
    id = "laskin_iv",
    type = "pkml",
    file = "laskin_1982_iv.pkml"
  )
)
removeObservedData(project, "laskin_iv")
```

```{r}
getObservedDataNames(project)
```

## Loading observed data

`loadObservedData()` resolves every declared source into a named list of `DataSet` objects, keyed by DataSet name. `createDataCombined()` (and therefore `createPlots()`) calls it internally to evaluate the observed side of a figure, so you rarely call it yourself.

```{r}
observed <- loadObservedData(project)
names(observed)
```

`getObservedDataNames()` returns the same names without reloading (it caches after the first load), which is how you discover what a project knows before referencing it in a DataCombined.

## Source type 1: Excel

An `excel` source names a workbook, an importer configuration, and the sheets to read. The bundled example's source is exactly this. To add another, point at a file under the project's data folder and supply the importer configuration and sheet names:

```{r, eval = FALSE}
# shown but not run: the example already declares this exact source, and a
# duplicate file id aborts
addObservedData(
  project,
  list(
    type = "excel",
    file = "Aciclovir_TimeValuesData.xlsx",
    importerConfiguration = "esqlabs_dataImporter_configuration.xml",
    sheets = "Laskin 1982.Group A"
  )
)
```

The importer configuration is an OSPS data-importer XML that describes how the columns map to time, value, and metadata; the `sheets` field lists which sheets to import. The file paths are relative to the project's data folder (`project$paths$dataFolder`).

## Source type 2: PKML

A `pkml` source reads a `DataSet` that was saved to a PKML file. Here we build a small `DataSet`, save it into the project's data folder as PKML, and declare it as a source:

```{r}
data_folder <- project$paths$dataFolder

pkml_data <- ospsuite::DataSet$new(name = "Aciclovir extra cohort")
pkml_data$setValues(xValues = c(0, 1, 2, 4, 8), yValues = c(0, 9, 6, 3, 1))
pkml_data$xDimension <- ospsuite::ospDimensions$Time
pkml_data$yDimension <- ospsuite::ospDimensions$`Concentration (molar)`
ospsuite::saveDataSetToPKML(pkml_data, file.path(data_folder, "extra_cohort.pkml"))

addObservedData(project, list(type = "pkml", file = "extra_cohort.pkml"))
getObservedDataNames(project)
```

The source definition is written to `definitions/observed-data/extra_cohort.pkml.json`; the PKML file stays under `Data/`, and the data is re-read from it whenever the project resolves observed data.

## Source type 3: Script

A `script` source names an R script that the package sources when the data is resolved. The script must return a `DataSet` or a list of them. Use it when the data is *derived* rather than fixed, so the recipe travels with the project and stays in step with its inputs. Write the script into the project (a `scripts/` subfolder of the data folder keeps it tidy), then declare it:

```{r}
script_dir <- file.path(data_folder, "scripts")
dir.create(script_dir, showWarnings = FALSE)

writeLines(
  c(
    'ds <- ospsuite::DataSet$new(name = "Digitized literature curve")',
    "ds$setValues(xValues = c(0, 1, 2, 4, 8), yValues = c(0, 11, 8, 4, 2))",
    "ds$xDimension <- ospsuite::ospDimensions$Time",
    "ds$yDimension <- ospsuite::ospDimensions$`Concentration (molar)`",
    "ds"
  ),
  file.path(script_dir, "literature_curve.R")
)

addObservedData(project, list(type = "script", file = "scripts/literature_curve.R"))
```

The id of this source is the file basename, `literature_curve.R`, while the DataSet it produces is named `Digitized literature curve`, which is the name a DataCombined references. Because the script runs each time the data is resolved, anything you can compute in R (digitized points, a transformation of a raw file, a join across several inputs) becomes reproducible observed data.

<div class="alert alert-warning">
**Security.** A `script` source runs the R file it names on your machine, with `source()`, whenever the data is resolved. Any R code in that file executes. Treat a project the way you would treat a script someone sends you: only load and resolve observed data from a project you trust. The first time a session sources a `script` source, the package warns you that this is happening.
</div>

## Source type 4: Programmatic

`addObservedData()` also accepts a `DataSet` you have already built in the session. Build it in R and pass it directly:

```{r}
scratch <- ospsuite::DataSet$new(name = "Scratch measurement")
scratch$setValues(xValues = c(0, 1, 3), yValues = c(0, 6, 2))

addObservedData(project, scratch)
```

Until you save, that `DataSet` lives on the in-memory project and the entry on disk is a `{type: "programmatic", name}` sentinel. On `saveProject()` the package writes the `DataSet` to a PKML file named after it (`Scratch measurement.pkml`) under the data folder and rewrites the entry to a `pkml` source. From then on it is an ordinary file-backed source: it survives a reload, and its remove id becomes the PKML file basename rather than the DataSet name.

Because saving turns a `programmatic` source into a `pkml` file, the data folder must be declared before you save; a project with no data folder cannot write the PKML and stops with a clear message. A `programmatic` sentinel that reaches a fresh session without its `DataSet` (a hand-authored declaration, or a project opened before it was ever saved) resolves to nothing, and the load warns you by name so it never vanishes silently.

## Remove a source

`removeObservedData()` takes the source id and accepts a vector to remove several at once:

```{r}
removeObservedData(project, c("extra_cohort.pkml", "Scratch measurement"))
getObservedDataNames(project)
```

If a removed source is still referenced by a DataCombined, the removal proceeds and leaves the reference dangling, which the next `validateProject()` (see `vignette("how-to-validate-a-project")`) reports as a problem, so prune the DataCombined too.

## Where to go next

To pair observed data with simulated curves and turn it into report figures, see `vignette("how-to-plot-results")`. To check that every source and its references hang together, see `vignette("how-to-validate-a-project")`.

For the required fields of each source type and the id rules, see `vignette("reference-definition-fields")`. For how to choose between the four types, and why a `script` source is the one that carries a trust cost, see `vignette("about-the-plotting-model")`.
