---
title: "How to run simulations"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How to run simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)

library(esqlabsR)
```

Once a project defines one or more scenarios, you run them with a single function, `runScenarios()`. A scenario is a named, runnable simulation specification (a model, an optional subject, the dosing, the parameters, the time grid, and the outputs to record); designing them is covered in `vignette("how-to-design-scenarios")`. Here we take a project that already has scenarios and focus on running them and handling the results.

This article uses the worked example project bundled with the package, so every chunk runs as written. Load it with `loadProject()`:

```{r load-project}
project <- loadProject(exampleProjectPath())

project$definitions$scenarios
```

The example defines three scenarios: an individual scenario, a population scenario, and an individual scenario that runs to steady state first.

## Running scenarios

`runScenarios()` takes the project and, optionally, the names of the scenarios to run. The first call in a session initializes PK-Sim, which takes a moment; subsequent calls are faster.

To run a single scenario, name it:

```{r run-one}
results <- runScenarios(project, scenarios = "aciclovir_iv")

names(results)
```

To run a subset, pass a vector of names:

```{r run-subset, eval = FALSE}
results <- runScenarios(
  project,
  scenarios = c("aciclovir_iv", "aciclovir_iv_steadystate")
)
```

To run every scenario in the project, leave `scenarios` as its default `NULL`:

```{r run-all, eval = FALSE}
results <- runScenarios(project)
```

In this article we keep the runnable example to the single individual scenario above, because each scenario builds and simulates a full PBPK model and that work adds up quickly.

## Get at the results

`runScenarios()` returns a named list keyed by scenario name. Each entry is one **Scenario Result**, a plain list bundling the prepared simulation, the raw OSPS results, the extracted output values, and the population. `vignette("reference-project-file")` describes all four fields.

```{r result-fields}
scenarioResult <- results[["aciclovir_iv"]]
```

For analysis or plotting, the field you want is almost always `outputValues`, which splits into the simulated `data` and its `metaData`:

```{r output-values}
head(scenarioResult$outputValues$data)
```

One point worth checking if a downstream function complains: a Scenario Result is **not** an OSPS `SimulationResults`. The `SimulationResults` is only the inner `results` field.

```{r results-class}
class(scenarioResult$results)
```

The `population` field tells you which kind of scenario you ran. It is `NULL` for an individual scenario, and holds the sampled `Population` for a population one, in which case `outputValues` carries one trajectory per virtual individual rather than a single curve.

```{r individual-population}
scenarioResult$population
```

## Save results and reload them later

Simulating a model is the slow step, so it is good practice to save results once and reload them for downstream work such as redrawing a figure, rather than re-simulating each time. `saveScenarioResults()` writes each Scenario Result to disk: the results as a `.csv`, the simulation as a `.pkml`, and, for population scenarios, the population as a `<scenario>_population.csv`.

By default the destination is derived from the project's output folder. You can also pass an explicit `outputFolder`; here we send the output to a temporary directory so nothing is written into the project tree:

```{r save-results}
resultsFolder <- withr::local_tempdir()

saveScenarioResults(
  results,
  project,
  outputFolder = resultsFolder
)
```

`loadScenarioResults()` reads the saved files back, given the scenario names and the folder they were written to. Passing the `project` restricts the reloaded output values to each scenario's declared output paths, so the reloaded column set matches what `runScenarios()` produced. The reload restores all four fields of each Scenario Result, including `population` for population scenarios:

```{r load-results}
reloaded <- loadScenarioResults(
  scenarios = "aciclovir_iv",
  resultsFolder = resultsFolder,
  project = project
)
```

The reloaded record has the same shape as the one `runScenarios()` returned, so any code that consumes a Scenario Result works on either.

## Tune the run

`runScenarios()` accepts a `simulationRunOptions` argument, an `ospsuite::SimulationRunOptions` object, controlling solver settings and whether scenarios are simulated in parallel. Leaving it at its default `NULL` uses the package defaults, which is right for most analyses. Pass one explicitly when you need to tune the run, for instance to spread many scenarios across several cores.

A project can also carry its own defaults in `defaultSimulationRunOptions`, which `runScenarios()` falls back on when you pass no options of your own; see `vignette("reference-project-file")` for the fields it accepts.

## Where to go next

With results in hand, turn them into figures with `vignette("how-to-plot-results")`. For the fields of a Scenario Result and the files `saveScenarioResults()` writes, see `vignette("reference-project-file")`.
