---
title: "How to customize sensitivity plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{How to customize sensitivity plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.showtext = TRUE
)
library(esqlabsR)
```

The three sensitivity plotting functions (`sensitivitySpiderPlot()`, `sensitivityTimeProfiles()`, and `sensitivityTornadoPlot()`) produce publication-ready figures by default. This article shows how to adjust their appearance: first with a `defaultPlotConfiguration` object, then with direct function arguments for one-off overrides, and finally with `{ggplot2}` and `{patchwork}` for anything the configuration cannot express.

The examples reuse the aciclovir sensitivity analysis from `vignette("how-to-run-a-sensitivity-analysis")`. See that article for the analysis setup; here we assume the resulting `analysis` object is in scope.

Reusing the analysis from `vignette("how-to-run-a-sensitivity-analysis")`:

```{r analysis}
project <- loadProject(exampleProjectPath())
simulation <- runScenarios(project, scenarios = "aciclovir_iv")[["aciclovir_iv"]]$simulation

outputPaths <- c(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
)
parameterPaths <- c(
  "Lipophilicity" = "Aciclovir|Lipophilicity",
  "GFR fraction" = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Glomerular Filtration-GFR-Aciclovir|GFR fraction"
)

analysis <- sensitivityCalculation(simulation, outputPaths, parameterPaths)
```

## Set the appearance with a configuration object

The `defaultPlotConfiguration` object is the central way to control the appearance of all three plots. When you pass none, each function builds one with `createEsqlabsPlotConfiguration()`, which is why the figures are house-styled by default.

Each function recognizes only a subset of the fields, because each has a different visual structure, and a field a function does not recognize has no effect. `vignette("reference-definition-fields")` has the support matrix showing which field works with which plot.

### Create and apply a configuration

Create a configuration object and set the fields you want to control:

```{r plot-config}
myConfiguration <- createEsqlabsPlotConfiguration()

myConfiguration$linesColor <- c("#4D8076", "#C34A36")
myConfiguration$title <- "Aciclovir - Peripheral Venous Blood"
myConfiguration$subtitle <- "Customized Plot"
myConfiguration$legendPosition <- "right"
```

Pass this object to any sensitivity plotting function:

```{r spider-plot-config, fig.height=4}
sensitivitySpiderPlot(
  sensitivityCalculation = analysis,
  defaultPlotConfiguration = myConfiguration
)
```

When a named vector is used for `parameterPaths`, those names are also used as labels in the plots.

### Override one setting for a single call

Some settings are also exposed as direct function arguments: `xAxisScale` and `yAxisScale` on `sensitivityTimeProfiles()`, `xAxisType` and `yAxisType` on `sensitivitySpiderPlot()`. A direct argument **overrides** the configuration object, which is how you make a one-off change without editing a configuration you reuse elsewhere.

Here the configuration asks for a log y-axis and the `yAxisScale = "lin"` argument wins:

```{r time-profile-override, fig.height=4}
myConfiguration <- createEsqlabsPlotConfiguration()

myConfiguration$linesColor <- c("#d8b365", "#5ab4ac")
myConfiguration$subtitle <- "Customized Plot"
myConfiguration$yAxisScale <- "log"

sensitivityTimeProfiles(
  sensitivityCalculation = analysis,
  yAxisScale = "lin",
  defaultPlotConfiguration = myConfiguration
)
```

### Change a plot after it is built

For anything the configuration cannot express, edit the returned object with `{ggplot2}` and `{patchwork}`. Each plotting function returns a **list of `patchwork` objects**, one entry per output path. Because a `patchwork` object is a composite of individual `ggplot2` plots, the `&` operator applies a change to every subplot at once.

```{r spider-plot-process, fig.height=4}
library(ggplot2)
library(patchwork)

spiderPlots <- sensitivitySpiderPlot(
  sensitivityCalculation = analysis,
  pkParameters = c("AUC_inf", "C_max")
)

# Extract the plot for a specific output path
spiderPlot <- spiderPlots[[outputPaths["Aciclovir_PVB"]]]

# Modify plot using `patchwork` and `ggplot2` functions
spiderPlot &
  theme(legend.position = "top") & # Place legend above the plot
  guides(col = guide_legend(ncol = 2)) & # Display legend entries in two columns
  scale_color_manual(values = c("#66a61e", "#d95f02")) & # Use custom colors for lines
  coord_cartesian(ylim = c(80, 125)) # Zoom in on the y-axis range
```

## Where to go next

For the analysis that produces these results, see `vignette("how-to-run-a-sensitivity-analysis")`. For which configuration field each plot recognizes, see `vignette("reference-definition-fields")`. For the full catalog of `PlotConfiguration` fields and their defaults, see the [TLF-Library reference](https://www.open-systems-pharmacology.org/TLF-Library/reference/PlotConfiguration.html) and its [plot-configuration article](https://www.open-systems-pharmacology.org/TLF-Library/articles/plot-configuration.html).
