---
title: "Observed data"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Observed data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)
library(esqlabsR)
```

Comparing a simulation against measured data is what turns a model run into a diagnostic. In `esqlabsR`, measured data is **Observed Data**: a named series of time/value points. It is the package's term for what OSPS implements as a `DataSet`. This article covers how observed data enters a project, the four ways to declare a source, and how the project resolves that data when you build a figure.

It assumes you already know what a Project is and how to load one (see `vignette("projects")`). For turning observed data and simulated curves into figures, see `vignette("plot-results")`, which owns the plotting model; this article owns how the data gets into the project in the first place.

## An observed-data source is a definition

A project does not store the measured numbers themselves. It stores a **source definition**: a small record under `definitions/observed-data/<id>.json` that says where the data comes from and how to read it. The actual `DataSet` is loaded on the fly from that definition whenever it is needed (at plot time, at validation time, or when you call `loadObservedData()` yourself), so the data is never baked into the project and always reflects the current source file.

There are four source types, distinguished only by where the data comes from:

- **`excel`** reads one or more sheets of an Excel workbook through an OSPS data-importer configuration.
- **`pkml`** reads a `DataSet` saved to a PKML file.
- **`script`** sources an R script that builds and returns a `DataSet` (or several). This is the reproducible programmatic path.
- **`programmatic`** holds a `DataSet` you built in the current session. It lives only in memory and is not reproducible across a reload (see the caveat below).

Throughout, scaffold a writable copy of the bundled example so the sources you add are written to a project you own, not the installed package:

```{r}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))
```

The example already ships with one observed-data source, an Excel source. You can see it as a definition on the project:

```{r}
project$definitions$observedData[[1]]
```

## A note on observed-data ids

Two strings attach to every source, and they are not always equal, so keep them distinct:

- **The source id** is the remove handle that `removeObservedData()` matches. It is taken verbatim from the source (not canonicalized like other definitions): the file basename for an `excel`, `pkml`, or `script` source (for example `Aciclovir_TimeValuesData.xlsx`), or the `DataSet` name for a `programmatic` source.
- **The DataSet name** is the reference handle that a DataCombined uses and `getObservedDataNames()` lists. It comes from the data itself: the Excel importer expands a sheet into a long, fully qualified name; a `pkml` or `script` source uses the `DataSet$name`.

Because the two can differ, read the DataSet name from `getObservedDataNames()` rather than typing it by hand.

```{r}
getObservedDataNames(project)
```

## Loading observed data

`loadObservedData()` resolves every declared source into a named list of `DataSet` objects, keyed by DataSet name. `createDataCombined()` (and therefore `createPlots()`) calls it internally to evaluate the observed side of a figure, so you rarely call it yourself.

```{r}
observed <- loadObservedData(project)
names(observed)
```

`getObservedDataNames()` returns the same names without reloading (it caches after the first load), which is how you discover what a project knows before referencing it in a DataCombined.

## Source type 1: Excel

An `excel` source names a workbook, an importer configuration, and the sheets to read. The bundled example's source is exactly this. To add another, point at a file under the project's data folder and supply the importer configuration and sheet names:

```{r, eval = FALSE}
# shown but not run: the example already declares this exact source, and a
# duplicate file id aborts
addObservedData(
  project,
  list(
    type = "excel",
    file = "Aciclovir_TimeValuesData.xlsx",
    importerConfiguration = "esqlabs_dataImporter_configuration.xml",
    sheets = "Laskin 1982.Group A"
  )
)
```

The importer configuration is an OSPS data-importer XML that describes how the columns map to time, value, and metadata; the `sheets` field lists which sheets to import. The file paths are relative to the project's data folder (`project$paths$dataFolder`).

## Source type 2: PKML

A `pkml` source reads a `DataSet` that was saved to a PKML file. Here we build a small `DataSet`, save it into the project's data folder as PKML, and declare it as a source:

```{r}
data_folder <- project$paths$dataFolder

pkml_data <- ospsuite::DataSet$new(name = "Aciclovir extra cohort")
pkml_data$setValues(xValues = c(0, 1, 2, 4, 8), yValues = c(0, 9, 6, 3, 1))
pkml_data$xDimension <- ospsuite::ospDimensions$Time
pkml_data$yDimension <- ospsuite::ospDimensions$`Concentration (molar)`
ospsuite::saveDataSetToPKML(pkml_data, file.path(data_folder, "extra_cohort.pkml"))

addObservedData(project, list(type = "pkml", file = "extra_cohort.pkml"))
getObservedDataNames(project)
```

The source definition is written to `definitions/observed-data/extra_cohort.pkml.json`; the PKML file stays under `Data/`, and the data is re-read from it whenever the project resolves observed data.

## Source type 3: Script

A `script` source names an R script that the package sources at load time. The script must return a `DataSet` or a list of `DataSet` objects. This is the most reproducible programmatic path: the data is regenerated from code every time, so it survives a reload and travels with the project. Write the script into the project (a `scripts/` subfolder of the data folder keeps it tidy), then declare it:

```{r}
script_dir <- file.path(data_folder, "scripts")
dir.create(script_dir, showWarnings = FALSE)

writeLines(
  c(
    'ds <- ospsuite::DataSet$new(name = "Digitized literature curve")',
    "ds$setValues(xValues = c(0, 1, 2, 4, 8), yValues = c(0, 11, 8, 4, 2))",
    "ds$xDimension <- ospsuite::ospDimensions$Time",
    "ds$yDimension <- ospsuite::ospDimensions$`Concentration (molar)`",
    "ds"
  ),
  file.path(script_dir, "literature_curve.R")
)

addObservedData(project, list(type = "script", file = "scripts/literature_curve.R"))
```

The id of this source is the file basename, `literature_curve.R`, while the DataSet it produces is named `Digitized literature curve` (the name a DataCombined references). Because the script runs each time the data is resolved, anything you can compute in R (digitized points, a transformation of a raw file, a join across several inputs) becomes reproducible observed data.

<div class="alert alert-warning">
**Security.** A `script` source runs the R file it names on your machine, with `source()`, whenever the data is resolved. Any R code in that file executes. Treat a project the way you would treat a script someone sends you: only load and resolve observed data from a project you trust. The first time a session sources a `script` source, the package warns you that this is happening.
</div>

## Source type 4: Programmatic

`addObservedData()` also accepts a `DataSet` you have already built in the session. The package keeps that `DataSet` on the in-memory project and writes only a `{type: "programmatic", name}` sentinel to disk, because a live `DataSet` is not JSON-serializable. Build the `DataSet` as in the PKML example, then pass it directly:

```{r}
scratch <- ospsuite::DataSet$new(name = "Scratch measurement")
scratch$setValues(xValues = c(0, 1, 3), yValues = c(0, 6, 2))

addObservedData(project, scratch)
```

This is convenient for a quick exploration, but be aware of the trade-off: the sentinel records the name, not the data, so a `programmatic` source is **not reproducible across a reload**. If you load the project in a fresh session, the source is declared but the `DataSet` behind it is gone. `addObservedData()` reminds you of this and recommends the `script` alternative, which is the reproducible way to express the same intent. Prefer a `script` source for anything you want to keep.

## Removing a source

`removeObservedData()` takes the source id (the remove handle described above: the file basename for an `excel`, `pkml`, or `script` source, the DataSet name for a `programmatic` one) and accepts a vector to remove several at once:

```{r}
removeObservedData(project, c("extra_cohort.pkml", "Scratch measurement"))
getObservedDataNames(project)
```

If a removed source is still referenced by a DataCombined, the removal proceeds and leaves the reference dangling, which the next `validateProject()` (see `vignette("validate-project")`) reports as a problem, so prune the DataCombined too.

## Where to go next

You now have all four ways observed data enters a project, and you know that the data is resolved on the fly from a source definition rather than stored in the project. To pair that observed data with simulated curves and turn it into report figures, see `vignette("plot-results")`. To check that every observed-data source and its references hang together, see `vignette("validate-project")`.
