---
title: "Parameter identification"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Parameter identification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.showtext = TRUE
)

library(esqlabsR)
```

```{css, echo = FALSE}
table {
  width: 100%;
  display: block;
  overflow-x: auto;
}
```

Parameter identification (PI) estimates model parameters by fitting simulated outputs to observed data. In `esqlabsR` a PI task lives inside a project, alongside its scenarios, output paths, and observed data, and is run with a single call. This article shows how to run the parameter identification that ships with the bundled example project and how to assemble one yourself with the programmatic authoring functions. It builds on the [`OSPSuite.ParameterIdentification`](https://www.open-systems-pharmacology.org/OSPSuite.ParameterIdentification/) package, which performs the actual optimization.

To follow along you only need a loaded project; see `vignette("projects")` for how a project is organized and loaded. Here we work entirely with the example project and never write anything into your project tree.

## The PI task model

A parameter identification is described by a **PI task**. A task is a plain-data record that names three things:

- the **scenarios** it fits (one or more named scenarios already defined in the project),
- a set of **PI Parameters**, one per quantity to estimate, each giving a model parameter path, the scenarios it applies to, a starting value, and lower and upper bounds, and
- a set of **PI Output Mappings**, each pairing one model output (an output path of the project) with one Observed Data series (see `vignette("observed-data")`), so the optimizer knows which simulated curve to compare against which measurements.

A task can also carry an optional configuration block (the optimization algorithm, the confidence-interval method, and similar settings). Tasks are stored on the project under `project$parameterIdentification`, keyed by task id.

## Running the predefined task

The bundled example project defines a single PI task, `aciclovirsimple`. It fits the parameter `EHC continuous fraction` on the scenario `aciclovir_iv`, mapping the simulated output `aciclovir_pvb` to an observed plasma series (from Laskin 1982, Group A).

Load the project and inspect the task:

```{r loadProject}
# Work on a writable copy of the bundled example: saving the PI tasks we add
# below writes to this project's own files rather than to the read-only example
# shipped with the package.
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))

project$parameterIdentification[["aciclovirsimple"]]
```

Run it with `runPI()`. By default `runPI()` runs every task on the project; pass `tasks` to restrict the run to specific tasks. The first call is slow because it initializes PK-Sim and builds the simulation before optimizing.

```{r runPredefined}
results <- runPI(project, tasks = "aciclovirsimple")
```

## The shape of the result

`runPI()` returns one entry per task. Each entry is a list with a `task` element (the parameter-identification object that was built and run) and a `result` element (the fit); if the optimization fails numerically, `result` is `NULL` and an `error` element carries the message while the run continues with the remaining tasks, whereas build errors (an unknown parameter path, output, or observed dataset) fail hard and stop the call so a typo surfaces immediately.

```{r resultShape}
fit <- results[["aciclovirsimple"]]
names(fit)
```

Print the result for a human-readable optimization summary:

```{r printResult}
fit$result
```

## Reading the estimates

Call `toDataFrame()` on the result to read the estimates as a tidy data frame, one row per estimated parameter, with the final `estimate` next to its `initialValue`, bounds, and uncertainty columns (`sd`, `cv`, `lowerCI`, `upperCI`):

```{r readEstimates}
fit$result$toDataFrame()
```

## When uncertainty cannot be quantified

A fit can converge and still report no usable uncertainty: when `sd`, `cv`, and the confidence interval all come back `NA`, the estimate is returned but its precision is unknown. This is what happens for `aciclovirsimple` above. Because a converged fit with no uncertainty is easy to mistake for a good one, `runPI()` emits a warning naming the task and parameter, with the likely causes:

- a singular or ill-conditioned Hessian,
- the estimate sitting at one of its bounds, or
- an objective function insensitive to the parameter.

When you see it, treat the estimate with caution and revisit the bounds, the output mapping, or whether the data constrain the parameter at all.

## Building a task programmatically

Instead of relying on a task stored in the project, you can author one in memory with the `add*` family. `addPITask()` creates a task and requires at least one PI Parameter and one PI Output Mapping at creation, so you seed it with existing records and grow it afterwards. The records are the same plain-data values the project already holds, so the simplest seed is to reuse the ones from the example task.

```{r addTask}
template <- project$parameterIdentification[["aciclovirsimple"]]

project <- addPITask(
  project,
  id = "ehc_fit",
  scenarios = "aciclovir_iv",
  parameters = template$parameters,
  outputMappings = template$outputMappings
)
```

`addPIParameter()` adds another parameter to fit. Give it the task id, the model parameter path, the scenarios it applies to, and the starting value with its bounds:

```{r addParameter}
project <- addPIParameter(
  project,
  task = "ehc_fit",
  path = "Organism|Liver|EHC continuous fraction",
  scenarios = "aciclovir_iv",
  minValue = 0.4,
  maxValue = 0.9,
  startValue = 0.6,
  id = "ehc_alt"
)
```

`addPIOutputMapping()` adds another output-to-data pairing, naming an output path already registered in the project and the Observed Data series to compare it against:

```{r addMapping}
observedId <- template$outputMappings[[1]]$observedDataId

project <- addPIOutputMapping(
  project,
  task = "ehc_fit",
  outputPath = "aciclovir_pvb",
  observedData = observedId,
  scenarios = "aciclovir_iv",
  scaling = "lin",
  id = "pvb_alt"
)
```

Once a task has the parameters and mappings you want, run it the same way as the predefined one with `runPI(project, tasks = "ehc_fit")`.

## Removing parameters and mappings

`removePIParameter()` and `removePIOutputMapping()` take the task id and the id of the record to drop. A task must always keep at least one parameter and one mapping, so removing the last of both leaves an empty task with nothing to fit. Rather than leave that empty shell on the project, `esqlabsR` removes the now-empty task for you and warns that it did so. The following strips `ehc_fit` down to nothing; the final removal triggers the auto-removal:

```{r removeAndAutoRemove}
project <- removePIParameter(project, "ehc_fit", "ehc_alt")
project <- removePIParameter(project, "ehc_fit", template$parameters[[1]]$id)
project <- removePIOutputMapping(project, "ehc_fit", "pvb_alt")
project <- removePIOutputMapping(project, "ehc_fit", template$outputMappings[[1]]$id)

project$parameterIdentification
```

The example task `aciclovirsimple` is untouched, since we only edited the in-memory copy of `ehc_fit`; nothing was written back to disk.

## Where to go next

The scenarios and output paths a PI task fits come from the rest of the project: see `vignette("design-scenarios")` for authoring scenarios and `vignette("observed-data")` for loading the measurements an output mapping compares against. For the underlying optimization, the confidence-interval methods, and the available algorithms, see the [`OSPSuite.ParameterIdentification`](https://www.open-systems-pharmacology.org/OSPSuite.ParameterIdentification/) package.
