---
title: "Plot results"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Plot results}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.showtext = TRUE
)

library(esqlabsR)
```

Clear, consistent figures are a core part of model diagnostics and of any client-facing report. This article shows how to turn scenario results into figures with `{esqlabsR}`: first the plots a project already defines, then how to author your own, attach observed data, apply the esqLABS house style, and export to a file.

Every figure below runs against the example project bundled with the package; loading it and running one scenario produces them all.

## The plotting model

`{esqlabsR}` describes figures with three building blocks that recur throughout this article.

- A **DataCombined** pairs one or more simulated curves with optional observed data. It is the data input for a chart: it says *which* scenario output and *which* observed data belong together, and it can carry per-curve transformations (offsets, scale factors).
- A **Plot** describes a single chart: its **Plot Type**, the **DataCombined** it draws from, and its axis, aggregation, and styling settings.
- A **Plot Grid** composes one or more **Plots** into one laid-out figure (a multi-panel arrangement). A **Plot** is usually rendered as a panel inside a **Plot Grid**, but can also be rendered on its own (see below).

A project stores all three as definitions under `definitions/data-combined/`, `definitions/plots/`, and `definitions/plot-grids/`, so the figures travel with the project. You can also assemble them yourself in R, which is covered below.

```{mermaid}
classDiagram
    direction LR
    class DataCombined {
        dataCombinedId
        simulated: label, scenario, path
        observed: label, dataSet
        ...
    }
    class Plot {
        plotId
        dataCombined
        plotType
        title
        ...
    }
    class PlotGrid {
        plotGridId
        plots
        title
        ...
    }
    DataCombined --> Plot
    Plot --> PlotGrid

    note for DataCombined "Simulation results paired with observed data"
    note for Plot "The data and type of a single chart"
    note for PlotGrid "Plots composed into one laid-out figure"
```

## Load the example and run a scenario

Every figure in this article is built from the results of one simulated scenario. Author against a writable copy of the example so the figures you add later do not change the installed package: `initProject()` scaffolds a fresh copy in a temporary directory, which you load and run the `aciclovir_iv` scenario against.

```{r setup-run}
project_dir <- withr::local_tempdir()
initProject(destination = project_dir, type = "example", createExcel = FALSE)
project <- loadProject(file.path(project_dir, "Project.json"))

scenarioResults <- runScenarios(
  project,
  scenarios = "aciclovir_iv"
)
```

The first `runScenarios()` call in a session initializes PK-Sim and is slow; later runs are fast. `scenarioResults` is keyed by scenario name; pass it to any plotting function below.

## Generate a project's predefined plots

The example project already defines a **DataCombined** (`aciclovir_individual`), a **Plot** (`p1`, of type `individual`), and a **Plot Grid** (`individual_diagnostics`). `createPlots()` reads those definitions and builds the figures, resolving each **DataCombined** against the scenario results you pass in.

```{r create-plots}
plotGrids <- createPlots(project, scenarioResults = scenarioResults)
```

Despite its name, `createPlots()` returns **Plot Grids**, not single **Plots**: the returned list is keyed by the project's **Plot Grid** names.

```{r show-predefined}
plotGrids$individual_diagnostics
```

To build only a subset of the project's grids, pass their names to `plotGrids`.

```{r create-plots-subset}
someGrids <- createPlots(
  project,
  plotGrids = "individual_diagnostics",
  scenarioResults = scenarioResults
)
```

To render a single **Plot** on its own, without wrapping it in a grid, pass its id to `plots`. The result is keyed by the plot's id, and the entry is a single `ggplot` object rather than a `patchwork` grid.

```{r create-plots-single}
single <- createPlots(
  project,
  plots = "p1",
  scenarioResults = scenarioResults
)

single$p1
```

`plotGrids` and `plots` are independent selectors, so you can request grids and standalone plots in the same call.

## Attach observed data

Comparing simulated curves against measurements is what makes a diagnostic figure useful. A project declares its observed data and resolves it on demand; how observed data enters a project (the four source types, the source definitions, and `loadObservedData()`) is covered in `vignette("observed-data")`. For plotting, you need just one handle: the **DataSet name** a **DataCombined** references, which `getObservedDataNames()` lists.

```{r observed}
getObservedDataNames(project)
```

Read this name from `getObservedDataNames()` rather than typing the long, fully qualified Excel name by hand. The predefined `aciclovir_individual` **DataCombined** already pairs the simulated curve with it, which is why the figure above overlays measurements on the prediction. Reference the same name when you author your own **DataCombined**.

## Author your own figures

Beyond a project's predefined figures, you can describe new ones in R with `addDataCombined()`, `addPlot()`, and `addPlotGrid()`. Like every authoring function, these edit the project in memory; `saveProject()` then writes each definition to the project's `definitions/data-combined/`, `definitions/plots/`, or `definitions/plot-grids/` folder. Because you scaffolded a writable copy above, a save lands in the temporary project, not the installed package.

First, define a **DataCombined** that pairs the simulated peripheral-venous-blood curve with the observed data. Curves that share a `group` are plotted together.

```{r add-datacombined}
observedName <- getObservedDataNames(project)[1]

addDataCombined(
  project,
  id = "aciclovir_report",
  simulated = list(list(
    label = "Aciclovir simulated",
    scenario = "aciclovir_iv",
    path = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)",
    group = "Aciclovir PVB"
  )),
  observed = list(list(
    label = "Aciclovir observed",
    dataSet = observedName,
    group = "Aciclovir PVB"
  ))
)
```

Next, describe a **Plot** that draws from this **DataCombined**. The `plotType` selects the kind of chart:

- `individual` plots a time profile for an individual scenario.
- `population` plots a time profile for a population scenario; its `quantiles` field (which applies only to population plots) controls the displayed percentile bands.
- `observedVsSimulated` plots simulated against observed values.
- `residualsVsSimulated` plots residuals against simulated values.
- `residualsVsTime` plots residuals against time.

Any field accepted by a plot configuration can be passed through `...`, for example a `title` or an axis unit.

```{r add-plot}
addPlot(
  project,
  id = "aciclovir_profile",
  dataCombined = "aciclovir_report",
  plotType = "individual",
  title = "Aciclovir IV, individual profile",
  xUnit = "h"
)
```

Finally, compose the **Plot** into a **Plot Grid**. A grid can hold a single plot or several; list every plot id you want in the figure.

```{r add-grid}
addPlotGrid(
  project,
  id = "report_figure",
  plots = "aciclovir_profile",
  title = "Aciclovir report figure"
)
```

Build the authored grid the same way as the predefined ones.

```{r build-authored}
reportGrids <- createPlots(
  project,
  plotGrids = "report_figure",
  scenarioResults = scenarioResults
)

reportGrids$report_figure
```

## Apply the esqLABS house style

`createPlots()` already styles figures with the esqLABS look, so the figures above are house-styled out of the box. When you build plots more directly, two helpers give you the same configurations to start from:

- `createEsqlabsPlotConfiguration()` returns a plot configuration carrying the esqLABS defaults for fonts, sizes, legend placement, and colors.
- `createEsqlabsPlotGridConfiguration()` returns the matching grid configuration (panel tags, title sizing, alignment).

```{r esqlabs-config}
plotConfig <- createEsqlabsPlotConfiguration()
gridConfig <- createEsqlabsPlotGridConfiguration()
```

The palette behind that look is `esqlabsColors()`, which extrapolates between the esqLABS blue, red, and green for any number of curves. Use it whenever you need house colors outside the plotting helpers.

```{r colors}
esqlabsColors(3)
```

## Export a finished figure

A **Plot Grid** is a standard `ggplot`/`patchwork` object, so save it with `ggplot2::ggsave()`. Choose the file extension to pick the format (`.png`, `.pdf`, `.svg`, and so on) and set the size and resolution to suit your report.

```{r ggsave}
outputDir <- withr::local_tempdir()

ggplot2::ggsave(
  filename = file.path(outputDir, "aciclovir_report.png"),
  plot = reportGrids$report_figure,
  width = 7,
  height = 5,
  dpi = 300
)
```

To export several figures, loop over the named list that `createPlots()` returns and write each grid to its own file, deriving the file name from the grid name.

```{r ggsave-loop}
for (name in names(reportGrids)) {
  ggplot2::ggsave(
    filename = file.path(outputDir, paste0(name, ".png")),
    plot = reportGrids[[name]]
  )
}
```

## Where to go next

The **Plot Type** charts shown here are produced by `{ospsuite}`'s plotting engine; its DataCombined articles cover the chart types and their customization in depth, and the OSPSuite-R observed-data article covers loading measurements from Excel or PKML. For building and running the scenarios that feed these figures, see `vignette("esqlabsR")` and `vignette("run-simulations")`.
