---
title: "Project configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project-structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)
library(esqlabsR)
```

## Project Configuration Overview

The project configuration is the central piece of any `esqlabsR` project. It defines the structure of your project, specifies where all necessary files are stored, and provides the foundation for all workflows in the package. The project configuration consists of:

- **Project structure**: A standardized folder hierarchy that organizes your models, data, parameters, scenarios, and results
- **ProjectConfiguration object**: An R object that stores paths to all project files and directories
- **Configuration files**: A JSON configuration file and supporting Excel files that define the project setup

## Initialize the Project Structure

Project structure can be initialized by calling the function `initProject()`.

```{r, eval=FALSE, message = FALSE}
library(esqlabsR)

initProject()
```

This will create a set of directories and files that match the expected project's 
structure.


```{r, echo = FALSE}
fs::dir_tree(path = esqlabsR:::projectDirectory("Example"), glob = "*.md|*.R", invert = TRUE)
```
## Create a ProjectConfiguration

The `ProjectConfiguration` stores the paths to the model-, data-, and other files
and directories. It will be used at several steps in `esqlabsR` workflows so it 
needs to be created before performing any simulations.

By printing the `ProjectConfiguration`, we can see the locations of all files 
used in the workflows:

```{r, eval = TRUE}
projectConfigPath <- system.file(
  "extdata", "projects", "Example", "ProjectConfiguration.json",
  package = "esqlabsR"
)
my_project_configuration <- loadProject(projectConfigPath)
```

Now that the project structure is initialized and the `ProjectConfiguration` is
created, read `vignette("design-scenarios")` to continue the process.
To learn more about `ProjectConfiguration`, read the following sections.

## Validating Project Configuration

Before running simulations, it is recommended to validate your project configuration
files to catch potential issues early. The `validateAllConfigurations()` function
performs comprehensive validation across all Excel configuration files and returns
structured results.

### Validation Process

The validation system checks your configuration files for:

- **Structural issues**: Missing required sheets or columns
- **Data integrity**: Empty sheets, invalid data types, required fields
- **Uniqueness constraints**: Duplicate IDs or names that must be unique
- **Cross-references**: Invalid references between configuration files (e.g., scenarios referencing non-existent individuals)

### Three-Tier Validation Results

Validation results are categorized into three levels:

1. **Critical Errors**: Blocking issues that prevent data import and must be fixed (e.g., missing required sheets, missing required columns)
2. **Warnings**: Non-blocking issues that allow import but indicate potential problems (e.g., empty optional sheets, data type mismatches)
3. **Valid Data**: Successfully validated data ready for use

### Usage Example

```{r, eval=FALSE}
# Validate all configuration files in a project
validation_results <- validateAllConfigurations(my_project_configuration)

# Or validate directly from the Excel file path
validation_results <- validateAllConfigurations(exampleProjectConfigurationPath())

# Check if there are any critical errors
if (isAnyCriticalErrors(validation_results)) {
  # Get detailed summary
  summary <- validationSummary(validation_results)
  print(paste("Critical errors found in:", paste(summary$files_with_errors, collapse = ", ")))
  print(paste("Total critical errors:", summary$total_critical_errors))
} else {
  print("All validations passed!")
}
```

### Interpreting Validation Results

The `validateAllConfigurations()` function returns a named list of `validationResult`
objects, one for each configuration file validated. Each result contains:

- `critical_errors`: List of blocking issues
- `warnings`: List of non-blocking issues
- `data`: The validated data (if validation succeeded)

You can access individual file results:

```{r, eval=FALSE}
# Check scenarios validation
scenarios_result <- validation_results$scenarios

if (scenarios_result$has_critical_errors()) {
  print("Scenarios file has critical errors:")
  print(scenarios_result$critical_errors)
}

if (length(scenarios_result$warnings) > 0) {
  print("Scenarios file has warnings:")
  print(scenarios_result$warnings)
}
```

The validation system ensures data quality before simulation, helping you identify
and fix configuration issues early in your workflow.

## Details

### Change ProjectConfiguration from R

If required, you can change the location of one of the files or folders using relative or absolute paths

```{r, eval=FALSE}
# change the location of the output folder
projectConfiguration$outputFolder <- "../anotherOutputFolder"

# change the location of the model parameters file
projectConfiguration$modelParametersFile <- "absolute/path/to/anotherModelParameters.xlsx"
```


### About `ProjectConfiguration.json`

The `ProjectConfiguration.json` file is the primary configuration file used by
`esqlabsR` to generate a `ProjectConfiguration` object which is the central
piece of a project. It should be located in the root folder of the project.

This file defines *where* all the necessary files are stored in the project
folder, which scenarios to run, and how observed data is loaded. Paths specified
in the JSON should be *relative* to the `ProjectConfiguration.json` location.

Supporting Excel files (e.g., `ModelParameters.xlsx`, `Individuals.xlsx`,
`Populations.xlsx`) provide detailed parametrization for scenarios. These files
have a specific purpose and template and are described in the next sections.

A `ProjectConfiguration.xlsx` file is also present for backwards compatibility
and can be used with `importProjectConfigurationFromExcel()` to generate or
update the JSON configuration.

### Version Control and Project Sharing

Since `ProjectConfiguration.json` is a text-based format, it works naturally
with version control systems like Git. This means you can:

- **Track changes** to your project configuration in version control systems like Git
- **Share projects** with team members easily
- **Create backups** before making significant changes
- **Ensure reproducibility** across different environments

#### Migrating from Excel to JSON

If you have an existing Excel-based project, `importProjectConfigurationFromExcel()`
converts your Excel configuration files into a single JSON file:

```{r, eval=FALSE}
# Migrate an Excel-based project to JSON
importProjectConfigurationFromExcel("ProjectConfiguration.xlsx")
```

By default, the JSON file will be created in the same directory as your
`ProjectConfiguration.xlsx` file with the name `ProjectConfiguration.json`.

#### Exporting back to Excel

The `exportProjectConfigurationToExcel()` function regenerates Excel configuration
files from a `ProjectConfiguration` object:

```{r, eval=FALSE}
# Load the project from JSON
pc <- loadProject("ProjectConfiguration.json")

# Export back to Excel files
exportProjectConfigurationToExcel(pc)
```

This function will:
1. Create the necessary directory structure
2. Regenerate all Excel configuration files

#### Workflow Examples

**Team Collaboration**: When working with a team:

1. **Initial Setup**: Load and commit the JSON project configuration
   ```{r, eval=FALSE}
pc <- loadProject("ProjectConfiguration.json")
   ```

2. **Version Control**: Commit the JSON file to your Git repository
   ```bash
   git add ProjectConfiguration.json
   git commit -m "Add project configuration"
   ```

3. **Team Sharing**: Team members clone the repo and load the project
   ```{r, eval=FALSE}
pc <- loadProject("ProjectConfiguration.json")
   ```

4. **Working with Excel**: If Excel editing is needed, export and re-import
   ```{r, eval=FALSE}
exportProjectConfigurationToExcel(pc)
# Edit Excel files...
importProjectConfigurationFromExcel("ProjectConfiguration.xlsx")
   ```

The version control features in `esqlabsR` provide a robust foundation for team collaboration and project management. By using JSON as the primary format, you can maintain project history in version control systems, share configurations easily with team members, create reliable backups before making changes, and ensure reproducibility across different environments.
