---
title: "Project configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project-structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)
library(esqlabsR)
```

## Project Configuration Overview

The project configuration is the central piece of any `esqlabsR` project. It defines the structure of your project, specifies where all necessary files are stored, and provides the foundation for all workflows in the package. The project configuration consists of:

- **Project structure**: A standardized folder hierarchy that organizes your models, data, parameters, scenarios, and results
- **ProjectConfiguration object**: An R object that stores paths to all project files and directories
- **Configuration files**: Excel files (or JSON snapshots) that define the project setup

## Initialize the Project Structure

Project structure can be initialized by calling the function `initProject()`.

```{r, eval=FALSE, message = FALSE}
library(esqlabsR)

initProject()
```

This will create a set of directories and files that match the expected project's 
structure.


```{r, echo = FALSE}
fs::dir_tree(path = esqlabsR:::exampleDirectory("TestProject"), glob = "*.md|*.R", invert = TRUE)
```
## Create a ProjectConfiguration

The `ProjectConfiguration` stores the paths to the model-, data-, and other files
and directories. It will be used at several steps in `esqlabsR` workflows so it 
needs to be created before performing any simulations.

By printing the `ProjectConfiguration`, we can see the locations of all files 
used in the workflows:

```{r, eval = TRUE}
my_project_configuration <- createProjectConfiguration(path = exampleProjectConfigurationPath())
```

Now that the project structure is initialized and the `ProjectConfiguration` is 
created, read `vignette("design-scenarios")` to continue the process.
To learn more about `ProjectConfiguration`, read the following sections.

## Details

### Change ProjectConfiguration from R

If required, you can change the location of one of the files or folders using relative or absolute paths

```{r, eval=FALSE}
# change the location of the output folder
projectConfiguration$outputFolder <- "../anotherOutputFolder"

# change the location of the model parameters file
projectConfiguration$modelParametersFile <- "absolute/path/to/anotherModelParameters.xlsx"
```


### About `ProjectConfiguration.xlsx`

The `ProjectConfiguration.xlsx` file will be used by `esqlabsR` to generate a
`ProjectConfiguration` object which is the central piece of a project. It should 
be located in the root folder of the project.

This file defines *where* all the necessary files are stored in the project 
folder. All the path specified in the `Value` column should be *relative* to the
`ProjectConfiguration.xlsx` location.

All these directories and files have a specific purpose and template and are 
describe in next sections.

```{r, echo = FALSE}
ProjectConfiguration_excel <- readxl::read_xlsx(exampleProjectConfigurationPath())

reactable::reactable(ProjectConfiguration_excel,
  columns = list(
    "Property" = reactable::colDef(minWidth = 50),
    "Value" = reactable::colDef(show = FALSE),
    "Description" = reactable::colDef(minWidth = 100)
  ),
  outlined = TRUE,
  pagination = FALSE,
  highlight = TRUE
)
```

### Version Control and Project Sharing

While Excel provides an excellent interface for quickly editing and creating new entries in your project configuration, it has limitations when it comes to version control and team collaboration - Excel files are binary and don't work well with Git, making it cumbersome to share changes between team members.

`esqlabsR` addresses these limitations by providing powerful version control and project sharing capabilities through its snapshot and restore functionality. This allows you to:

- **Track changes** to your project configuration in version control systems like Git
- **Share projects** with team members easily
- **Create backups** before making significant changes
- **Ensure reproducibility** across different environments

#### Creating Project Snapshots

The `snapshotProjectConfiguration()` function exports all Excel configuration files in your project to a single JSON file. This JSON format is ideal for version control as it's:

- **Text-based**: Easy to track changes in Git
- **Human-readable**: Can be reviewed and compared
- **Complete**: Contains all project configuration data

```{r, eval=FALSE}
# Create a snapshot from a ProjectConfiguration object
my_project_configuration <- createProjectConfiguration("ProjectConfiguration.xlsx")
snapshotProjectConfiguration(my_project_configuration)

# Or create a snapshot directly from the Excel file path
snapshotProjectConfiguration("ProjectConfiguration.xlsx")
```

By default, the JSON file will be created in the same directory as your `ProjectConfiguration.xlsx` file with the name `ProjectConfiguration.json`.

#### Restoring from Snapshots

The `restoreProjectConfiguration()` function recreates your entire project configuration from a JSON snapshot:

```{r, eval=FALSE}
# Restore a project from a JSON snapshot
restored_config <- restoreProjectConfiguration("ProjectConfiguration.json")
```

This function will:
1. Create the necessary directory structure
2. Regenerate all Excel configuration files
3. Return a new `ProjectConfiguration` object

#### Checking Project Status

Use `projectConfigurationStatus()` to check if your Excel files are synchronized with your JSON snapshot:

```{r, eval=FALSE}
# Check if files are in sync
status <- projectConfigurationStatus("ProjectConfiguration.xlsx", "ProjectConfiguration.json")
print(status$in_sync) # TRUE if synchronized, FALSE otherwise
```

#### Workflow Examples

**Team Collaboration**: When working with a team:

1. **Initial Setup**: Create a snapshot of your project configuration
   ```{r, eval=FALSE}
snapshotProjectConfiguration("ProjectConfiguration.xlsx")
   ```

2. **Version Control**: Commit the JSON file to your Git repository
   ```bash
   git add ProjectConfiguration.json
   git commit -m "Add project configuration snapshot"
   ```

3. **Team Sharing**: Team members can restore the project
   ```{r, eval=FALSE}
restored_config <- restoreProjectConfiguration("ProjectConfiguration.json")
   ```

**Backup Strategy**: Before making significant changes to your project:

```{r, eval=FALSE}
# Create a backup snapshot
snapshotProjectConfiguration("ProjectConfiguration.xlsx")

# Make your changes...

# If something goes wrong, restore from the snapshot
restored_config <- restoreProjectConfiguration("ProjectConfiguration.json")
```

The version control features in `esqlabsR` provide a robust foundation for team collaboration and project management. By using snapshots, you can maintain project history in version control systems, share configurations easily with team members, create reliable backups before making changes, and ensure reproducibility across different environments.
