---
title: "Project configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project-structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)
library(esqlabsR)
```

## Project Configuration Overview

The project configuration is the central piece of any `esqlabsR` project. It defines the structure of your project, specifies where all necessary files are stored, and provides the foundation for all workflows in the package. The project configuration consists of:

- **Project structure**: A standardized folder hierarchy that organizes your models, data, parameters, scenarios, and results
- **Project object**: An R object that stores paths to all project files and directories
- **Configuration files**: A JSON configuration file and supporting Excel files that define the project setup

## Initialize the Project Structure

Project structure can be initialized by calling the function `initProject()`.

```{r, eval=FALSE, message = FALSE}
library(esqlabsR)

initProject()
```

This will create a set of directories and files that match the expected project's 
structure.


```{r, echo = FALSE}
fs::dir_tree(path = dirname(exampleProjectPath()), glob = "*.md|*.R", invert = TRUE)
```
## Create a Project

The `Project` stores the paths to the model-, data-, and other files
and directories. It will be used at several steps in `esqlabsR` workflows so it
needs to be created before performing any simulations.

By printing the `Project`, we can see the locations of all files
used in the workflows:

```{r, eval = TRUE}
projectConfigPath <- system.file(
  "extdata", "projects", "Example", "Project.json",
  package = "esqlabsR"
)
my_project <- loadProject(projectConfigPath)
my_project
```

Now that the project structure is initialized and the `Project` is
created, read `vignette("design-scenarios")` to continue the process.
To learn more about `Project`, read the following sections.

## Validating Project

Before running simulations, it is recommended to validate your project
configuration to catch potential issues early. The `validateProject()`
function accepts either a loaded `Project` object or a path to a `Project.json`
file, performs comprehensive validation across all project sections, and
returns structured results.

### Validation Process

The validation system checks your project for:

- **Required fields**: Missing required fields in individuals, populations, scenarios
- **Data integrity**: Invalid data types, values out of range
- **Uniqueness constraints**: Duplicate IDs or names that must be unique
- **Cross-references**: Invalid references between sections (e.g., scenarios referencing non-existent individuals or populations)

### Three-Tier Validation Results

Validation results are categorized into three levels:

1. **Critical Errors**: Blocking issues that prevent data import and must be fixed (e.g., missing required sheets, missing required columns)
2. **Warnings**: Non-blocking issues that allow import but indicate potential problems (e.g., empty optional sheets, data type mismatches)
3. **Valid Data**: Successfully validated data ready for use

### Usage Example

```{r, eval=FALSE}
# Validate a loaded project
validation_results <- validateProject(my_project)

# Or validate directly from a Project.json path
validation_results <- validateProject("Project.json")

# Check if there are any critical errors
if (isAnyCriticalErrors(validation_results)) {
  # Get detailed summary
  summary <- validationSummary(validation_results)
  print(paste("Critical errors found in:", paste(summary$files_with_errors, collapse = ", ")))
  print(paste("Total critical errors:", summary$total_critical_errors))
} else {
  print("All validations passed!")
}
```

### Interpreting Validation Results

The `validateProject()` function returns a named list of `validationResult`
objects, one for each project section validated. Each result contains:

- `critical_errors`: List of blocking issues
- `warnings`: List of non-blocking issues
- `data`: The validated data (if validation succeeded)

You can access individual section results:

```{r, eval=FALSE}
# Check scenarios validation
scenarios_result <- validation_results$scenarios

if (scenarios_result$has_critical_errors()) {
  print("Scenarios section has critical errors:")
  print(scenarios_result$critical_errors)
}

if (length(scenarios_result$warnings) > 0) {
  print("Scenarios section has warnings:")
  print(scenarios_result$warnings)
}

# Available sections: individuals, populations, scenarios, outputPaths,
# modelParameters, applications, plots, crossReferences
```

The validation system ensures data quality before simulation, helping you identify
and fix configuration issues early in your workflow.

## Details

### Change Project from R

If required, you can change the location of one of the files or folders using relative or absolute paths

```{r, eval=FALSE}
# change the location of the output folder
project$outputFolder <- "../anotherOutputFolder"

# change the location of the model parameters file
project$modelParamsFile <- "absolute/path/to/anotherModelParameters.xlsx"
```


### About `Project.json`

The `Project.json` file is the primary configuration file used by
`esqlabsR` to generate a `Project` object which is the central
piece of a project. It should be located in the root folder of the project.

This file defines *where* all the necessary files are stored in the project
folder, which scenarios to run, and how observed data is loaded. Paths specified
in the JSON should be *relative* to the `Project.json` location.

Excel files (`ModelParameters.xlsx`, `Individuals.xlsx`, `Populations.xlsx`,
`Applications.xlsx`, `Scenarios.xlsx`, `Plots.xlsx`) are an **optional**
editing surface for the same content that lives in `Project.json`. Edit them,
then sync back to JSON with `importProjectFromExcel()`. `Project.json` remains
the authoritative store.

A `Project.xlsx` file can be generated from any JSON project via
`exportProjectToExcel()`; it is optional and not committed by default.
`importProjectFromExcel()` performs the reverse direction — generating or
updating `Project.json` from an Excel-based project.

### Observed Data

The `observedData` section in `Project.json` declares data sources that are
loaded via `loadObservedData()`. Each entry specifies a source type and the
information needed to load it.

#### Source Types

**Excel files** — loaded via a data importer configuration:

```json
{
  "type": "excel",
  "file": "StudyData.xlsx",
  "importerConfiguration": "importer_config.xml",
  "sheets": ["Group A", "Group B"]
}
```

**PKML files** — loaded directly:

```json
{
  "type": "pkml",
  "file": "observed_data.pkml"
}
```

**R scripts** — sourced at load time, must return a `DataSet` or list of `DataSet` objects:
```json
{
  "type": "script",
  "file": "scripts/generate_data.R"
}
```

All file paths are relative to `dataFolder`.

#### Loading Observed Data

```{r, eval=FALSE}
project <- loadProject("Project.json")

# Get all DataSet names (loads data on first call, then cached)
getObservedDataNames(project)

# Load all DataSets
dataSets <- loadObservedData(project)
```

#### Adding Data Programmatically

You can add observed data entries at runtime using `addObservedData()`:

```{r, eval=FALSE}
# Add a DataSet directly (uses dataSet$name as key)
my_data <- ospsuite::DataSet$new(name = "MyPKData")
my_data$setValues(xValues = c(0, 1, 2, 4), yValues = c(0, 10, 8, 5))
project$addObservedData(my_data)

# Add a config entry (e.g., a script to be loaded later)
project$addObservedData(list(
  type = "script",
  file = "scripts/additional_data.R"
))
```

When adding a `DataSet` directly, duplicate names are checked against all
existing DataSet names (including those from Excel, PKML, and script sources).

### Version Control and Project Sharing

Since `Project.json` is a text-based format, it works naturally
with version control systems like Git. This means you can:

- **Track changes** to your project configuration in version control systems like Git
- **Share projects** with team members easily
- **Create backups** before making significant changes
- **Ensure reproducibility** across different environments

#### Migrating from Excel to JSON

If you have an existing Excel-based project, `importProjectFromExcel()`
converts your Excel configuration files into a single JSON file:

```{r, eval=FALSE}
# Migrate an Excel-based project to JSON
importProjectFromExcel("Project.xlsx")
```

By default, `importProjectFromExcel()` writes the JSON output next to the
input file, using the same basename with `.xlsx` replaced by `.json` — so
`Project.xlsx` yields `Project.json` in the same directory. Pass the
`outputDir` argument to write elsewhere.

#### Exporting back to Excel

The `exportProjectToExcel()` function regenerates Excel configuration
files from a `Project` object:

```{r, eval=FALSE}
# Load the project from JSON
pc <- loadProject("Project.json")

# Export back to Excel files
exportProjectToExcel(pc)
```

This function will:
1. Create the necessary directory structure
2. Regenerate all Excel configuration files

> **Known gaps in the Excel round-trip:** Some `Project.json` content is not
> reproduced in Excel form — species-level parameters and other internal
> fields stay JSON-only. An export → edit-in-Excel → re-import round-trip
> preserves the user-facing configuration but may drop internal metadata.
> Prefer editing `Project.json` directly when fidelity matters.

#### Workflow Examples

**Team Collaboration**: When working with a team:

1. **Initial Setup**: Load and commit the JSON project configuration
   ```{r, eval=FALSE}
pc <- loadProject("Project.json")
   ```

2. **Version Control**: Commit the JSON file to your Git repository
   ```bash
   git add Project.json
   git commit -m "Add project configuration"
   ```

3. **Team Sharing**: Team members clone the repo and load the project
   ```{r, eval=FALSE}
pc <- loadProject("Project.json")
   ```

4. **Working with Excel**: If Excel editing is needed, export and re-import
   ```{r, eval=FALSE}
exportProjectToExcel(pc)
# Edit Excel files...
importProjectFromExcel("Project.xlsx")
   ```

The version control features in `esqlabsR` provide a robust foundation for team collaboration and project management. By using JSON as the primary format, you can maintain project history in version control systems, share configurations easily with team members, create reliable backups before making changes, and ensure reproducibility across different environments.
