---
title: "3. Run Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{run-simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)

library(esqlabsR)

projectConfigPath <- system.file(
  "extdata", "projects", "Example", "ProjectConfiguration.json",
  package = "esqlabsR"
)
projectConfiguration <- loadProject(projectConfigPath)
```

Running scenarios with `{esqlabsR}` is straightforward: load the project
configuration and call `runScenarios()`.

### Loading the project

A `ProjectConfiguration` object holds all information about the project,
including scenarios, individuals, populations, model parameters, and
observed data. It is created by loading a `ProjectConfiguration.json`
file with `loadProject()`:

```{r, eval=FALSE}
projectConfiguration <- loadProject("path/to/ProjectConfiguration.json")
```

### Running scenarios

The `runScenarios()` function accepts a `ProjectConfiguration` object
and runs the scenarios defined in it. You can optionally specify which
scenarios to run by name. The output is a named list, where the names
are scenario names, and the values are lists containing the simulation,
`SimulationResults`, the output values, and the population (if the
scenario is a population simulation).

```{r runScenarios, results='hold'}
simulatedScenariosResults <- runScenarios(
  projectConfiguration,
  scenarioNames = "Aciclovir_iv"
)

# Each simulation is stored separately inside the simulatedScenariosResults
names(simulatedScenariosResults)

# Each simulation can be accessed using its name
simulatedScenariosResults$Aciclovir_iv$simulation

# Of course, it contains simulated results as dataframe
head(simulatedScenariosResults$Aciclovir_iv$outputValues$data)

# It also contains dataframe's metadata
head(simulatedScenariosResults$Aciclovir_iv$outputValues$metaData)
```

During the model development/fitting phase, you might want to test
parameter values before storing them in the Excel files.
You can pass custom parameters directly to `runScenarios()`:

```{r, eval=FALSE}
results <- runScenarios(
  projectConfiguration,
  scenarioNames = "Aciclovir_iv",
  customParams = customParams
)
```

It is a good idea to store simulation results as `*.csv` along with
simulations as `.*pkml` and, optionally, the applied population as
`*.csv` file and load them for further processing to avoid re-simulating
every time, e.g., a change to a figure showing the simulation results is
required. The convenient function for saving the results of a scenario
run is `exportScenarioResults`. Then, you will be able to restore the
results using `importScenarioResults`.

Now that the results have been generated, you can proceed to the next
step in the `vignette("plot-results")`.
<!-- Read the following sections to learn more about the options for running simulations. -->
