---
title: "Standard workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = "../tests/data/TestProject/Code/"
)
```

```{r, echo = FALSE, results = "hide", message = F}
library(esqlabsR)
```

### Defining scenarios

Within the `esqlabsR` framework, the simulations are run by defining and
executing multiple *scenarios*. A scenario is defined by the simulation
file containing the model structure, parametrization of the model,
application protocol, and (optionally) the physiology of the simulated
individual or population. To simplify scenarios set up, all these
information are stored in excel files with defined structure.

Information about location of scenario definition files, simulation
files, output folders etc. is stored in excel file
`ProjectConfiguration.xlsx` located in the folder `Code`. Usually it is
not required to edit the contents of this file.

The first step in the workflow is to create a `ProjectConfiguration`
from this file that will be used by the scripts to know where to find
the required data (for all following examples, we assume that the
current working directory is the `Code` folder):

```{r}
projectConfiguration <- createDefaultProjectConfiguration()
print(projectConfiguration)
```

Next, we will define a *scenario* that we want to simulate in the excel
file `Parameters/Scenarios.xlsx` (or other as set in
`projectConfiguration$scenarioDefinitionFile`). The package includes [an
example
scenario](https://github.com/esqLABS/esqlabsR/blob/HEAD/tests/data/TestProject/Parameters/Scenarios.xlsx)
that models the administration of a single dose of 250 mg aciclovir
intravenously to an individual with a 90 ml/min estimated glomerular
filtration rate. To define a scenario, create a new row in the
`Scenarios.xlsx` file with the following content:

-   **Scenario_name**: unique name of the scenario. The name must be a
    [valid R variables
    name](https://cran.r-project.org/doc/FAQ/R-FAQ.html#What-are-valid-names_003f).
-   **IndividualId**: Name (ID) of an individual. This name refers to
    the name of the individual as used in excel files
    `IndividualPhysiology.xlsx` for definition of the biometric
    properties of the simulated individual and
    `IndividualParameters.xlsx` for definition of individual-specific
    model parameters. **IndividualId** may be empty. In this case,
    standard individual as defined in the `pkml` simulation without any
    individual-specific model parameters will be simulated. 
To create an individual with specific biometric characteristics, define
an entry with the same individual id in the file
`IndividualPhysiology.xlsx`. The same individual can be used in multiple
scenarios. Note that though it is technically possible to apply a
species different from the one in the original simulation, the
correctness of the results is not guaranteed as there exist some
structural differences between the species.

-   **ModelParameterSheets**: A list of sheet names from the
    `ModelParameter.xlsx` file, separated by a `,`. Each sheet must have
    the columns `Container Path`, `Parameter Name`, `Value`, and
    `Units`. Parameter values from specified sheets will be applied to
    the model in the order of their definition. E.g., if we define
    `Global, Aciclovir`, then parameters from the `Aciclovir` sheet will
    be applied after the `Global` parameters. **ModelParameterSheets**
    may be empty or specify as many sheets as required. Note that the
    specified sheets must be present in the `ModelParameter.xlsx` file.
    The purpose of this approach is to have *global* parameters that can
    be applied to most scenarios, and separate set of parameters for
    e.g. different *disease states* (parameter sheets `Healthy` and
    `CKD`), or separate parametrization of different *compounds* (sheet
    `Aciclovir`).  
We can further refine the parametrization of the scenario by specifying
the individual parameters in the `IndividualParameters.xlsx` file.
Create a sheet with the name as the *IndividualId* specified for the
respective scenario and populate it the same structure as the
`ModelParameters.xlsx` file. This way we can define e.g.
individual-specific clearance values or, as in our case, the glomerular
filtration rate. Individual specific parameters are applied after the
parameters defined in the *ModelParameterSheets*. We can use an
individual in multiple scenarios. If an individual is specified in the
scenario definition, but no sheet with this name exists in the
`IndividualParameters.xlsx` file, this step is simply ignored.

-   **ApplicationProtocol**: name of the application protocol that will
    be applied. Applications are defined as a set of parameters that
    will be applied to the model in the file
    `ApplicationParameters.xlsx`. For each application, create a sheet
    with the name as specified in the **ApplicationProtocol** entry and
    populate it with the same structure as the `ModelParameter.xlsx`
    file...
  Configuring application protocols this way requires that the loaded
simulation includes all possible applications that can be turned on and
off by setting parameters, e.g. the `Dose` or `Start time` parameters.
As it might be cumbersome to manually create entries for all
administration parameters, we can use the
`getAllApplicationParameters()` function to get a list of all (constant)
parameters located in the `Applications` container. In the following
example, we will extract application parameters for the molecule
`Aciclovir` from the example simulation:
```{r}
sim <- loadSimulation(system.file("extdata", "Aciclovir.pkml", package = "ospsuite"))
applicationParams <- getAllApplicationParameters(sim)
print(applicationParams)
```
and export the parameters to an excel file using the
`exportParametersToXLS()` function:
```{r, eval=FALSE}
exportParametersToXLS(parameters = applicationParams, paramsXLSpath = "../Applications.xlsx")
```
The created excel file will have the same structure as all
Parameter-files and can be directly loaded in MoBi or R using the
`readParametersFromXLS()` function.

- **SimulationTime**: Time of the simulation. The [output intervel](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/setOutputInterval.html) 
of the simulation will be set to the range [0 - `SimulationTime`] with the resolution 
of 1 point per minute.

- **SimulationTimeUnit**: Unit for `SimulationTime`. For supported units, see 
`ospsuite::ospUnits`.

- **SteadyState**: If `TRUE`, the model will be simulated for a "sufficiently long"
time (1000 minutes by default). The steady-state time can be specified in the 
`ScenarioConfiguration` (see Section [Running scenarios](#running-scenarios)).

- **ModelFile**: Name of the `pkml` file with the simulation. Must be located in 
the folder defined in `ProjectConfiguration$modelFolder`.

### Running scenarios

Scenarios are then executed by creating a `ScenarioConfiguration` object
and calling a `runScenarios()` function with a specific scenario name
(`TestScenario` in this case):

```{r runScenarios}
scenarioConfiguration <- ScenarioConfiguration$new(projectConfiguration)
OutputPaths <- enum(list(
  Aciclovir_PVB = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
))
simulatedScenarios <- runScenarios(
  scenarioNames = c("TestScenario"),
  scenarioConfiguration = scenarioConfiguration,
  customParams = NULL, saveSimulationsToPKML = FALSE
)
```

The `runScenarios()` function relies on an `OutputPaths` variable being
defined in the current environment. The variable is expected to contain
path(s) to outputs generated by the model. In our case this is the
simulated concentration of Aciclovir in the peripheral venous blood
plasma.

The `runScenarios()` function returns a named list of lists with
`Simulation` objects, `SimulationResults` objects, and vectors of output
values.

```{r simulatedScenarios}
print(simulatedScenarios$TestScenario$results)
```

```{r outputValues}
print(head(simulatedScenarios$TestScenario$outputValues$data))
```

### Troubleshooting

-   The `runScenarios()` function will fail if the `OutputPaths`
    variable is missing in the current environment. Make sure to have
    this variable set up immediately before calling the `runScenarios()`
    function.

More detailed information on function signatures can be found in:

-   `esqlabsR` documentation on:
    -   [ProjectConfiguration
        class](https://esqlabs.github.io/esqlabsR/reference/ProjectConfiguration.html)
    -   [ScenarioConfiguration
        class](https://esqlabs.github.io/esqlabsR/reference/ScenarioConfiguration.html)
    -   [readScenarioConfigurationFromExcel()](https://esqlabs.github.io/esqlabsR/reference/readScenarioConfigurationFromExcel.html)
    -   [runScenarios()](https://esqlabs.github.io/esqlabsR/reference/runScenarios.html)
-   `ospsuite` documentation on:
    -   [Simulation
        class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/Simulation.html)
    -   [SimulationResults
        class](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/SimulationResults.html)
